---
title: "07-Regression"
output:
  html_document:
    toc: yes
  html_notebook: default
  pdf_document:
    toc: yes
---

```{r message=FALSE, warning=FALSE, echo=F, eval=TRUE,paged.print = FALSE}
options(digits = 3)
```

# Regression 

This chapter is primarily based on: 

* Field, A., Miles J., & Field, Z. (2012): Discovering Statistics Using R. Sage Publications (**chapters 6, 7, 8**).
* James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013): An Introduction to Statistical Learning with Applications in R, Springer (**chapter 3**)

## Correlation

Before we start with regression analysis, we will review the basic concept of correlation first. Correlation helps us to determine the degree to which the variation in one variable, X, is related to the variation in another variable, Y. 

### Correlation coefficient

The correlation coefficient summarizes the strength of the linear relationship between two metric (interval or ratio scaled) variables. Let's consider a simple example. Say you conduct a survey to investigate the relationship between the attitude towards a city and the duration of residency. The "Attitude" variable can take values between 1 (very unfavourable) and 12 (very favorable), and the "duration of residency" is measured in years. Let's further assume for this example that the attitude measurement represents an interval scale (although it is usually not realistic to assume that the scale points on an itemized rating scale have the same distance). To keep it simple, let's further assume that you only asked 12 people. We can create a short dataset like this:    

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,paged.print = FALSE}
library(psych)
attitude <- c(6,9,8,3,10,4,5,2,11,9,10,2)
duration <- c(10,12,12,4,12,6,8,2,18,9,17,2)
att_data <- data.frame(attitude, duration)
att_data <- att_data[order(-attitude), ]
att_data$respodentID <- c(1:12)
str(att_data)
psych::describe(att_data[, c("attitude","duration")])
att_data
```

Let's look at the data. The following graph shows the individual data points for the "duration of residency"" variable, where the blue horizontal line represents the mean of the variable (`r round(mean(att_data$duration),2)`) and the vertical lines show the distance of the individual data points from the mean.

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Scores for duration of residency variable"}
library(ggplot2)
h <- round(mean(att_data$duration), 2)
ggplot(att_data, aes(x = respodentID, y = duration)) + 
  geom_point(size = 3, color = "deepskyblue4") + 
  scale_x_continuous(breaks = 1:12) + 
  geom_hline(data = att_data, aes(yintercept = mean(duration)), color ="deepskyblue4") +
  labs(x = "Observations",y = "Duration of residency", size = 11) +
  coord_cartesian(ylim = c(0, 18)) +
  geom_segment(aes(x = respodentID,y = duration, xend = respodentID, 
                   yend = mean(duration)), color = "deepskyblue4", size = 1) + 
  theme(axis.title = element_text(size = 16),
        axis.text  = element_text(size=16),
        strip.text.x = element_text(size = 16),
        legend.position="none") + 
  theme_bw()
```

You can see that there are some respondents that have been living in the city longer than average and some respondents that have been living in the city shorter than average. Let's do the same for the second variable ("Attitude"): 

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Scores for attitude variable"}
ggplot(att_data, aes(x = respodentID, y = attitude)) + 
  geom_point(size = 3, color = "#f9756d") + 
  scale_x_continuous(breaks = 1:12) + 
  geom_hline(data = att_data, aes(yintercept = mean(attitude)), color = "#f9756d") +
  labs(x = "Observations",y = "Attitude", size = 11) +
  coord_cartesian(ylim = c(0,18)) +
  geom_segment(aes(x = respodentID, y = attitude, xend = respodentID, 
                   yend = mean(attitude)), color = "#f9756d", size = 1) + 
  theme_bw()
```

Again, we can see that some respondents have an above average attitude towards the city (more favourable) and some respondents have a below average attitude towards the city. Let's plot the data in one graph now to see if there is some co-movement: 

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE,fig.align="center", fig.cap = "Scores for attitude and duration of residency variables"}
ggplot(att_data) + 
  geom_point(size = 3, aes(respodentID, attitude), color = "#f9756d") +
  geom_point(size = 3, aes(respodentID, duration), color = "deepskyblue4")  + 
  scale_x_continuous(breaks = 1:12) + 
  geom_hline(data = att_data, aes(yintercept = mean(duration)), color = "deepskyblue4") +
  geom_hline(data = att_data, aes(yintercept = mean(attitude)), color = "#f9756d") +
  labs(x = "Observations", y = "Duration/Attitude", size = 11) +
  coord_cartesian(ylim = c(0, 18)) +
  scale_color_manual(values = c("#f9756d", "deepskyblue4")) +
  theme_bw()
```

We can see that there is indeed some co-movement here. The variables <b>covary</b> because respondents who have an above (below) average attitude towards the city also appear to have been living in the city for an above (below) average amount of time and vice versa. Correlation helps us to quantify this relationship. Before you proceed to compute the correlation coefficient, you should first look at the data. We usually use a scatterpot to visualize the relationship between two metric variables:

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Scatterplot for durationand attitute variables"}
ggplot(att_data) + 
  geom_point(size = 3, aes(duration, attitude)) + 
  labs(x = "Duration",y = "Attitude", size = 11) +
  theme_bw()
```

How can we compute the correlation coefficient? Remember that the variance measures the average deviation from the mean of a variable:

\begin{equation} 
\begin{split}
s_x^2&=\frac{\sum_{i=1}^{N} (X_i-\overline{X})^2}{N-1} \\
     &= \frac{\sum_{i=1}^{N} (X_i-\overline{X})*(X_i-\overline{X})}{N-1}
\end{split}
(\#eq:variance)
\end{equation} 

When we consider two variables, we multiply the deviation for one variable by the respective deviation for the second variable: 

<p style="text-align:center;">
$(X_i-\overline{X})*(Y_i-\overline{Y})$
</p>

This is called the cross-product deviation. Then we sum the cross-product deviations:

<p style="text-align:center;">
$\sum_{i=1}^{N}(X_i-\overline{X})*(Y_i-\overline{Y})$
</p>

... and compute the average of the sum of all cross-product deviations to get the <b>covariance</b>:

\begin{equation} 
Cov(x, y) =\frac{\sum_{i=1}^{N}(X_i-\overline{X})*(Y_i-\overline{Y})}{N-1}
(\#eq:covariance)
\end{equation} 

You can easily compute the covariance manually as follows

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
x <- att_data$duration
x_bar <- mean(att_data$duration)
y <- att_data$attitude
y_bar <- mean(att_data$attitude)
N <- nrow(att_data)
cov <- (sum((x - x_bar)*(y - y_bar))) / (N - 1)
cov
```

Or you simply use the built-in ```cov()``` function:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
cov(att_data$duration, att_data$attitude)          # apply the cov function 
```

A positive covariance indicates that as one variable deviates from the mean, the other variable deviates in the same direction. A negative covariance indicates that as one variable deviates from the mean (e.g., increases), the other variable deviates in the opposite direction (e.g., decreases).

However, the size of the covariance depends on the scale of measurement. Larger scale units will lead to larger covariance. To overcome the problem of dependence on measurement scale, we need to convert convariance to a standard set of units through standardisation by dividing the covariance by the standard deviation (i.e., similar to how we compute z-scores).

With two variables, there are two standard deviations. We simply multiply the two standard deviations. We then divide the covariance by the product of the two standard deviations to get the standardised covariance, which is known as a correlation coefficient r:

\begin{equation} 
r=\frac{Cov_{xy}}{s_x*s_y}
(\#eq:corcoeff)
\end{equation} 

This is known as the product moment correlation (r) and it is straight-forward to compute:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
x_sd <- sd(att_data$duration)
y_sd <- sd(att_data$attitude)
r <- cov/(x_sd*y_sd)
r
```

Or you could just use the ```cor()``` function:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
cor(att_data[, c("attitude", "duration")], method = "pearson", use = "complete")
```

Properties of r:

* ranges from -1 to + 1
* +1 indicates perfect linear relationship
* -1 indicates perfect negative relationship
* 0 indicates no linear relationship
* ± .1 represents small effect
* ± .3 represents medium effect
* ± .5 represents large effect

### Significance testing

How can we determine if our two variables are significantly related? To test this, we denote the population moment correlation &rho;. Then we test the null of no relationship between variables:

$$H_0:\rho=0$$
$$H_1:\rho\ne0$$

The test statistic is: 

\begin{equation} 
t=\frac{r*\sqrt{N-2}}{\sqrt{1-r^2}}
(\#eq:cortest)
\end{equation} 

It has a t distribution with n - 2 degrees of freedom. Then, we follow the usual procedure of calculating the test statistic and comparing the test statistic to the critical value of the underlying probability distribution. If the calculated test statstic is larger than the critical value, the null hypothesis of no relationship between X and Y is rejected. 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
t_calc <- r*sqrt(N - 2)/sqrt(1 - r^2) #calculated test statistic
t_calc
df <- (N - 2) #degrees of freedom
t_crit <- qt(0.975, df) #critical value
t_crit
pt(q = t_calc, df = df, lower.tail = F) * 2 #p-value 
```

Or you can simply use the ```cor.test()``` function, which also produces the 95% confidence interval:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
cor.test(att_data$attitude, att_data$duration, alternative = "two.sided", method = "pearson", conf.level = 0.95)
```

To determine the linear relationship between variables, the data only needs to be measured using interval scales. If you want to test the significance of the association, the sampling distribution needs to be normally distributed (we usually assume this when our data are normally distributed or when N is large). If parametric assumptions are violated, you should use non-parametric tests:

* Spearman's correlation coefficient: requires ordinal data and ranks the data before applying Pearson's equation.
* Kendall's tau: use when N is small or the number of tied ranks is large.

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
cor.test(att_data$attitude, att_data$duration, alternative = "two.sided", method = "spearman", conf.level = 0.95)
cor.test(att_data$attitude, att_data$duration, alternative = "two.sided", method = "kendall", conf.level = 0.95)
```

Report the results:

A Pearson product-moment correlation coefficient was computed to assess the relationship between the duration of residence in a city and the attitude toward the city. There was a positive correlation between the two variables, r = 0.936, n = 12, p < 0.05. A scatterplot summarizes the results (Figure XY).

**A note on the interpretation of correlation coefficients:**

Correlation coefficients give no indication of the direction of causality. In our example, we can conclude that the attitude toward the city is more positive as the years of residence increases. However, we cannot say that the years of residence cause the attitudes to be more positive. There are two main reasons for caution when interpreting correlations:

* Third-variable problem: there may be other unobserved factors that affect the results.
* Direction of causality: Correlations say nothing about which variable causes the other to change (reverse causality: attitudes may just as well cause the years of residence variable).

## Regression

Correlations measure relationships between variables (i.e., how much two variables covary). Using regression analysis we can predict the outcome of a dependent variable (Y) from one or more independent variables (X). E.g., how many products will we sell if we increase the advertising expenditures by 1000 Euros? In regression analysis, we fit a model to our data and use it to predict the values of the dependent variable from one predictor variable (bivariate regression) or several predictor variables (multiple regression). The following table shows a comparision of correlation and regression analysis:

<br>

&nbsp; | Correlation	 | Regression	
-------------|--------------------------  | -------------------------- 
Estimated coefficient  | Coefficient of correlation (bounded between -1 and +1) | Regression coefficient (not bounded a priori)
Interpretation  | Linear association between two variables; Association is bidirectional | (Linear) relation between one or more independent variables and dependent variable; Relation is directional
Role of theory | Theory neither required nor testable  | Theory required and testable

<br>

### Simple linear regression

In simple linear regression, we assess the relationship between one dependent (regressand) and one independent (regressor) variable. The goal is to fit a line through a scatterplot of observations in order to find the line that best describes the data (scatterplot).

Suppose you are a marketing research analyst at a music label and your task is to suggest, on the basis of past data, a marketing plan for the next year that will maximize product sales. The data set that is available to you includes information on the sales of music downloads (thousands of units), advertising expenditures (in Euros), the number of radio plays an artist received per week (airplay), the nuber of previous releases of an artist (starpower), repertoire origin (country; 0 = local, 1 = international), and genre (1 = rock, 2 = pop, 3 = electronic). Let's load and inspect the data first: 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
regression <- read.table("https://raw.githubusercontent.com/IMSMWU/Teaching/master/MRDA2017/music_sales_regression.dat", 
                          sep = "\t", 
                          header = TRUE) #read in data
regression$country <- factor(regression$country, levels = c(0:1), labels = c("local", "international")) #convert grouping variable to factor
regression$genre <- factor(regression$genre, levels = c(1:3), labels = c("rock", "pop","electronic")) #convert grouping variable to factor
head(regression)
```

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, paged.print = FALSE}
psych::describe(regression) #descriptive statistics using psych
```

As stated above, regression analysis may be used to relate a quantitative response ("dependent variable") to one or more predictor variables ("independent variables"). In a simple linear regression, we have one dependent and one independent variable.  

Here are a few important questions that we might seek to address based on the data:

* Is there a relationship between advertising budget and sales? 
* How strong is the relationship between advertising budget and sales?
* Which other variables contribute to sales?
* How accurately can we estimate the effect of each variable on sales?
* How accurately can we predict future sales?
* Is the relationship linear?
* Is there synergy among the advertising activities?

We may use linear regression to answer these quesitons. Let's start with the first question and investigate the effect of advertising on sales. 

#### Estimating the coefficients

A simple linear regression model only has one predictor and can be written as:

\begin{equation} 
Y=\beta_0+\beta_1X+\epsilon
(\#eq:regequ)
\end{equation} 

In our specific context, let's consider only the influence of advertising on sales for now:

\begin{equation} 
Sales=\beta_0+\beta_1*adspend+\epsilon
(\#eq:regequadv)
\end{equation} 

The word "adspend" represents data on advertising expenditures that we have observed and &beta;<sub>1</sub> (the "slope"") represents the unknown relationship between advertising expenditures and sales. It tells you by how much sales will increase for an additional Euro spent on advertising. &beta;<sub>0</sub> (the "intercept") is the number of sales we would expect if no money is spent on advertising. Together, &beta;<sub>0</sub> and &beta;<sub>1</sub> represent the model coefficients or parameters. The error term (&epsilon;) captures everything that we miss by using our model, including, (1) misspecifications (the true relationship might not be linear), (2) omitted variables (other variables might drive sales), and (3) measurement error (our measurement of the variables might be imperfect).      

Once we have used our training data to produce estimates for the model coefficients, we can predict future sales on the basis of a particular value of advertising expenditures by computing:

\begin{equation} 
\hat{Sales}=\hat{\beta_0}+\hat{\beta_1}*adspend
(\#eq:predreg)
\end{equation} 

We use the hat symbol, <sup>^</sup>, to denote the estimated value for an unknown parameter or coefficient, or to denote the predicted value of the response (sales). In practice, &beta;<sub>0</sub> and &beta;<sub>1</sub> are unknown and must be estimated from the data to make predictions. In the case of our advertising example, the data set consists of the advertising budget and product sales (n = 200). Our goal is to obtain coefficient estimates such that the linear model fits the available data well. In other words, we fit a line through the scatterplot of observations and try to find the line that best describes the data. The following graph shows the scatterplot for our data, where the black line shows the regression line. The grey vertical lines shows the difference between the predicted values (the regression line) and the observed values. This difference is referred to as the residuals ("e").

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Ordinary least squares (OLS)"}
library(dplyr)
options(scipen = 999)
rm(regression)
rm(advertising)
regression <- read.table("https://raw.githubusercontent.com/IMSMWU/Teaching/master/MRDA2017/multiple_regression.dat", 
                          sep = "\t", 
                          header = TRUE) #read in data
lm <- lm(sales ~ adspend, data = regression)
regression$yhat <- predict(lm)

ggplot(regression, aes(x = adspend,y = sales)) + 
  geom_point(size = 2, color = "deepskyblue4") + 
  labs(x = "Advertising expenditure (in Euros)", y = "Sales", size = 11) +
  geom_segment(aes(x = adspend, y = sales, xend = adspend, yend = yhat), 
               color = "grey",size = 0.5, linetype = "solid", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme(axis.title = element_text(size = 16),
        axis.text  = element_text(size = 16),
        strip.text.x = element_text(size = 16),
        legend.position="none") + 
  theme_bw()
```

Estimation of the regression function is based on the idea of the method of least squares (OLS = ordinary least squares). The first step is to calculate the residuals by subtracting the observed values from the predicted values.

<p style="text-align:center;">
$e_i = Y_i-(\beta_0+\beta_1X_i)$
</p>

This difference is then minimized by minimizing the sum of the squared residuals:

\begin{equation} 
\sum_{i=1}^{N} e_i^2= \sum_{i=1}^{N} [Y_i-(\beta_0+\beta_1X_i)]^2\rightarrow min!
(\#eq:rss)
\end{equation} 

e<sub>i</sub>: Residuals (i = 1,2,...,N)<br>
Y<sub>i</sub>: Values of the dependent variable (i = 1,2,...,N) <br>
&beta;<sub>0</sub>: Intercept<br>
&beta;<sub>1</sub>: Regression coefficient / slope parameters<br>
X<sub>ni</sub>: Values of the nth independent variables and the ith observation<br>
N: Number of observations<br>

This is also referred to as the <b>residual sum of squares (RSS)</b>. Now we need to choose the values for &beta;<sub>0</sub> and &beta;<sub>1</sub> that minimize RSS. So how can we derive these values for the regression coefficient? The equation for &beta;<sub>1</sub> is given by:

\begin{equation} 
\hat{\beta_1}=\frac{COV_{XY}}{s_x^2}
(\#eq:slope)
\end{equation} 

The exact mathematical derivation of this formula is beyond the scope of this script, but the intuition is to calculate the first derivative of the squared residuals with respect to &beta;<sub>1</sub> and set it to zero, thereby finding the &beta;<sub>1</sub> that minimizes the term. Using the above formula, you can easily compute &beta;<sub>1</sub> using the following code:

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE}
cov_y_x <- cov(regression$adspend, regression$sales)
cov_y_x
var_x <- var(regression$adspend)
var_x
beta_1 <- cov_y_x/var_x
beta_1
```

The interpretation of &beta;<sub>1</sub> is as follows: 

For every extra Euros spent on advertising, sales can be expected to increase by `r round(beta_1,3)` units. Or, in other words, if we increase our marketing budget by 1,000 Euros, sales can be expected to increase by `r round(beta_1,3)*1000` units.

Using the estimated coefficient for &beta;<sub>1</sub> , it is easy to compute &beta;<sub>0</sub> (the intercept) as follows:

\begin{equation} 
\hat{\beta_0}=\overline{Y}-\hat{\beta_1}\overline{X}
(\#eq:intercept)
\end{equation} 

The R code for this is:

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE}
beta_0 <- mean(regression$sales) - beta_1*mean(regression$adspend)
beta_0
```

The interpretation of &beta;<sub>0</sub> is as follows: 

If we spend no money on advertising, we would expect to sell `r round(beta_0,3)` units.

You may also verify this based on a scatterplot of the data. The following plot shows the scatterplot including the regression line, which is estimated using OLS.  

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center", fig.cap = "Scatterplot"}
ggplot(regression, mapping = aes(adspend, sales)) + 
  geom_point(shape = 1) +
  geom_smooth(method = "lm", fill = "blue", alpha = 0.1) + 
  labs(x = "Advertising expenditures (EUR)", y = "Number of sales", colour = "store") + 
  theme_bw()
```

You can see that the regression line intersects with the y-axis at `r round(beta_0,2)`, which corresponds to the expected sales level when advertising expenditure (on the x-axis) is zero (i.e., the intercept &beta;<sub>0</sub>). The slope coefficient (&beta;<sub>1</sub>) tells you by how much sales (on the y-axis) would increase if advertising expenditures (on the x-axis) are increased by one unit.   

#### Significance testing

In a next step, we assess if the effect of advertsing on sales is statistically significant. This means that we test the null hypothesis H<sub>0</sub>: "There is no relationship between advertising and sales" versus the alternative hypothesis H<sub>1</sub>: "The is some relationship between advertising and sales". Or, to state this mathematically:

$$H_0:\beta_1=0$$
$$H_1:\beta_1\ne0$$

How can we test if the effect is statistically significant? Recall the generalized equation to derive a test statstic:

\begin{equation} 
test\ statistic = \frac{effect}{error}
(\#eq:teststatgeneral)
\end{equation} 

The effect is given by the &beta;<sub>1</sub> coefficient in this case. To compute the test statisitc, we need to come up with a measure of uncertainty around this estimate (the error). This is because we use information from a sample to estimate the least squares line to make inferences regarding the regression line in the entire population. Since we only have access to one sample, the regression line will be slightly different every time we take a different sample from the population. This is sampling variation and it is perfectly normal! It just means that we need to take into account the untercainty around the estimate, which is achieved by the standard error. Thus, the test statistic for our hypothesis is given by:

\begin{equation} 
t = \frac{\hat{\beta_1}}{SE(\hat{\beta_1})}
(\#eq:teststatreg)
\end{equation} 

After calculating the test statstic, we compare its value to the values that we would expect to find if there was no effect based on the t-distribution. In a regressin context, the degrees of freedom are given by ```N - p - 1``` where N is the sample size and p is the number of predictors. In our case, we have 200 observations and one predictor. Thus, the degrees of freedom is 200 - 1 - 1 = 198. In the regression output below, R provides the exact probability of observing a t value of this magnitude (or larger) if the null hypothesis was true. This probability is the p-value. A small p-value indicates that it is unlikely to observe such a substantial association between the predictor and the outcome variable due to chance in the absence of any real association between the predictor and the outcome.

To estimate the regression model in R, you can use the ```lm()``` function. Within the function, you first specify the dependent variable ("sales") and independent variable ("adspend") separated by a ```~``` (tilde). As mentioned previously, this is known as _formula notation_ in R. The ```data = regression``` argument specifies that the variables come from the data frame named "regression". Strictly speaking, you use the ```lm()``` function to create an object called "simple_regression," which holds the regression output. You can then view the results using the ```summary()``` function: 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
simple_regression <- lm(sales ~ adspend, data = regression) #estimate linear model
summary(simple_regression) #summary of results
```

Note that the estimated coefficients for &beta;<sub>0</sub> (`r round(summary(simple_regression)$coefficients[1],3)`) and &beta;<sub>1</sub> (`r round(summary(simple_regression)$coefficients[2],3)`) correspond to the results of our manual computation above. The associated t-values and p-values are given in the output. The t-values are larger than the critical t-values for the 95% confidence level, since the associated p-values are smaller than 0.05. In case of the coefficient for &beta;<sub>1</sub> this means that the probability of an association between the advertising and sales of the observed magnitude (or larger) is smaller than 0.05, if the value of &beta;<sub>1</sub> was, in fact, 0.     

The coefficients associated with the respective variables represent <b>point estimates</b>. To get a better feeling for the range of values that the coefficients could take, it is helpful to compute <b>confidence intervals</b>. A 95% confidence interval is defined as a range of values such that with a 95% probability, the range will contain the true unknown value of the parameter. For example, for &beta;<sub>1</sub>, the confidence interval can be computed as.

\begin{equation} 
CI = \hat{\beta_1}\pm(t_{1-\frac{\alpha}{2}}*SE(\beta_1))
(\#eq:regCI)
\end{equation} 

It is easy to compute confidence intervals in R using the ```confint()``` function. You just have to provide the name of you estimated model as an argument:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
confint(simple_regression)
```

For our model, the 95% confidence interval for &beta;<sub>0</sub> is [`r round(confint(simple_regression)[1,1],2)`,`r round(confint(simple_regression)[1,2],2)`], and the 95% confidence interval for &beta;<sub>1</sub> is [`r round(confint(simple_regression)[2,1],2)`,`r round(confint(simple_regression)[2,2],2)`]. Thus, we can conclude that when we do not spend any money on advertising, sales will be somewhere between `r round(confint(simple_regression)[1,1],0)` and `r round(confint(simple_regression)[1,2],0)` units on average. In addition, for each increase in advertising expenditures by one Euro, there will be an average increase in sales of between `r round(confint(simple_regression)[2,1],2)` and `r round(confint(simple_regression)[2,2],2)`. 

#### Assessing model fit

Once we have rejected the null hypothesis in favor of the alternative hypothesis, the next step is to investigate to what extent the model represents ("fits") the data. How can we assess the model fit?

* First, we calculate the fit of the most basic model (i.e., the mean)
* Then, we calculate the fit of the best model (i.e., the regression model)
* A good model should fit the data significantly better than the basic model
* R<sup>2</sup>: Represents the percentage of the variation in the outcome that can be explained by the model
* The F-ratio measures how much the model has improved the prediction of the outcome compared to the level of inaccuracy in the model

Similar to ANOVA, the calculation of model fit statistics relies on estimating the different sum of squares values. SS<sub>T</sub> is the difference between the observed data and the mean value of Y (aka. total variation). In the absence of any other information, the mean value of Y represents the best guess on where an observation at a given level of advertising will fall:

\begin{equation} 
SS_T= \sum_{i=1}^{N} (Y_i-\overline{Y})^2
(\#eq:regSST)
\end{equation} 

The following graph shows the total sum of squares:

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Total sum of squares"}
library(dplyr)
options(scipen = 999)
rm(regression)
rm(advertising)
regression <- read.table("https://raw.githubusercontent.com/IMSMWU/Teaching/master/MRDA2017/music_sales_regression.dat", 
                          sep = "\t", 
                          header = TRUE) #read in data
lm <- lm(sales ~ adspend, data = regression)

regression$yhat <- predict(lm)

ggplot(regression, aes(x = adspend, y = sales)) + 
  geom_point(size = 2, color = "deepskyblue4") + 
  labs(x = "Advertising", y = "Sales", size = 11) +
  geom_segment(aes(x = adspend, y = sales, xend = adspend, 
                   yend = mean(sales)), color = "grey", 
               size = 0.5, linetype = "solid", alpha = 0.8) + 
  geom_hline(data = regression, aes(yintercept = mean(sales)), color = "black", size = 1) +
  theme(axis.title = element_text(size = 16),
        axis.text  = element_text(size = 16),
        strip.text.x = element_text(size = 16),
        legend.position="none") + 
  theme_bw()
```

Based on our linear model, the best guess about the sales level at a given level of advertising is the predicted value. The model sum of squares (SS<sub>M</sub>) has the mathematical representation:

\begin{equation} 
SS_M= \sum_{j=1}^{c}  n_j(\overline{Y}_j-\overline{Y})^2
(\#eq:regSSM)
\end{equation} 

The model sum of squares represents the improvement in prediction resulting from using the regression model rather than the mean of the data. The following graph shows the model sum of squares for our example:

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Ordinary least squares (OLS)"}
ggplot(regression, aes(x = adspend, y = sales)) + 
  geom_point(size = 2, color = "deepskyblue4") + 
  labs(x = "Advertising",y = "Sales", size = 11) +
  geom_segment(aes(x = adspend, y = yhat, xend = adspend, yend = mean(sales)), 
               color = "grey", size = 0.5, linetype = "solid", alpha = 0.8) + 
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  geom_hline(data = regression, aes(yintercept = mean(sales)), color = "black", size = 1) +
  theme(axis.title = element_text(size = 16),
        axis.text  = element_text(size = 16),
        strip.text.x = element_text(size = 16),
        legend.position = "none") + 
  theme_bw()
```

The residual sum of squares (SS<sub>R</sub>) is the difference between the observed data and the predicted values along the regression line (i.e., the variation not explained by the model)

\begin{equation} 
SS_R= \sum_{j=1}^{c} \sum_{i=1}^{n} ({Y}_{ij}-\overline{Y}_{j})^2
(\#eq:regSSR)
\end{equation} 

The following graph shows the residual sum of squares for our example:

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.cap = "Ordinary least squares (OLS)"}
ggplot(regression, aes(x = adspend, y = sales)) + 
  geom_point(size = 2, color = "deepskyblue4") + 
  labs(x = "Advertising",y = "Sales", size = 11) +
  geom_segment(aes(x = adspend, y = sales, xend = adspend, yend = yhat),
               color = "grey", size = 0.5, linetype = "solid", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme(axis.title = element_text(size = 16),
        axis.text  = element_text(size = 16),
        strip.text.x = element_text(size = 16),
        legend.position = "none") + 
  theme_bw()
```

##### R-squared {-}

The R<sup>2</sup> statistic represents the proportion of variance that is explained by the model and is computed as:

\begin{equation} 
R^2= \frac{SS_M}{SS_T}
(\#eq:regSSR)
\end{equation} 

It takes values between 0 (very bad fit) and 1 (very good fit). Note that when the goal of your model is to predict future outcomes, a "too good" model fit can pose severe challanges. The reason is that the model might fit your specific sample so well, that it will only predict well within the sample but not generalize to other samples. This is called **overfitting** and it shows that there is a trade-off between model fit and out-of-sample predictive ability of the model, if the goal is to predict beyond the sample.

You can get a first impression of the fit of the model by inspecting the scatter plot as can be seen in the plot below. If the observations are highly dispersed around the regression line (left plot), the fit will be lower compared to a data set where the values are less dispersed (right plot).

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.height = 4, fig.width = 10, fig.cap="Good vs. bad model fit"}
library(cowplot)
library(gridExtra)
library(grid)
set.seed(44)
#x3 <- rlnorm(250, log(1), log(0.6))
options(scipen = 999)
options(digits = 2)
x1 <- rnorm(200,614.41,485)
x <- as.data.frame(subset(x1,x1>0))
names(x)<-c('adspend')
error <- rnorm(nrow(x))
sales <- round(134 + 0.094*x$adspend + 50*error)
sales_data <- data.frame(sales,x$adspend) 
names(sales_data)<-c('sales','adspend')
examplereg <- subset(sales_data,sales>0 & adspend>0)
#summary(examplereg)
lm <- lm(sales ~ adspend, data = examplereg)
#summary(lm)
examplereg$yhat <- predict(lm)

scatter_plot1 <- ggplot(examplereg,aes(adspend,sales)) +  
  geom_point(size=2,shape=1) +    # Use hollow circles
  geom_smooth(method="lm") + # Add linear examplereg line (by default includes 95% confidence region);
  scale_x_continuous(name="advertising expenditures", limits=c(0, 1800)) +
  scale_y_continuous(name="sales", limits=c(0, 400)) +
  theme_bw() +
  labs(title = paste0("R-squared: ",round(summary(lm)$r.squared,2))) 

#x3 <- rlnorm(250, log(1), log(0.6))
options(scipen = 999)
options(digits = 2)
x1 <- rnorm(200,614.41,485)
x <- as.data.frame(subset(x1,x1>0))
names(x)<-c('adspend')
error <- rnorm(nrow(x))
sales <- round(134 + 0.094*x$adspend + 20*error)
sales_data <- data.frame(sales,x$adspend) 
names(sales_data)<-c('sales','adspend')
#summary(sales_data)
examplereg <- subset(sales_data,sales>0 & adspend>0)
#summary(examplereg)

lm <- lm(sales ~ adspend, data = examplereg)
examplereg$yhat <- predict(lm)

scatter_plot2 <- ggplot(examplereg,aes(adspend,sales)) +  
  geom_point(size=2,shape=1) +    # Use hollow circles
  geom_smooth(method="lm") + # Add linear examplereg line (by default includes 95% confidence region);
  scale_x_continuous(name="advertising expenditures", limits=c(0, 1800)) +
  scale_y_continuous(name="sales", limits=c(0, 400)) +
  theme_bw() +
  labs(title = paste0("R-squared: ",round(summary(lm)$r.squared,2))) 

#p <- plot_grid(plot1, plot2, ncol = 2)
p <- plot_grid(scatter_plot1,scatter_plot2, ncol = 2)
print(p)
# now add the title
#title <- ggdraw() + draw_label("", fontface='bold')
#plot_full <- plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
#print(plot_full)
```

The R<sup>2</sup> statistic is reported in the regression output (see above). However, you could also extract the relevant sum of squares statistics from the regression object using the ```anova()``` function to compute it manually: 

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE,paged.print = FALSE}
anova(simple_regression) #anova results
```

Now we can compute R<sup>2</sup> in the same way that we have computed Eta<sup>2</sup> in the last section:

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE,paged.print = FALSE}
r2 <- anova(simple_regression)$'Sum Sq'[1]/(anova(simple_regression)$'Sum Sq'[1] + anova(simple_regression)$'Sum Sq'[2]) #compute R2
```

##### Adjusted R-squared {-}

Due to the way the R<sup>2</sup> statistic is calculated, it will never decrease if a new explanatory variable is introduced into the model. This means that every new independent variable either doesn't change the R<sup>2</sup> or increases it, even if there is no real relationship between the new variable and the dependent variable. Hence, one could be tempted to just add as many variables as possible to increase the R<sup>2</sup> and thus obtain a "better" model. However, this actually only leads to more noise and therefore a worse model. 

To account for this, there exists a test statistic closely related to the R<sup>2</sup>, the **adjusted R<sup>2</sup>**. It can be calculated as follows:

\begin{equation} 
\overline{R^2} = 1 - (1 - R^2)\frac{n-1}{n - k - 1}
(\#eq:adjustedR2)
\end{equation} 

where ```n``` is the total number of observations and ```k``` is the total number of explanatory variables. The adjusted adjusted R<sup>2</sup> is equal to or less than the regular R<sup>2</sup> and can be negative. It will only increase if the added variable adds more explanatory power than one expect by pure chance. Essentially, it contains a "penalty" for including unnecessary variables and therefore favors more parsimonious models. As such, it is a measure of suitability, good for comparing different models and is very useful in the model selection stage of a project. In R, the standard ```lm()``` function automatically also reports the adjusted R<sup>2</sup>.

##### F-test {-}

Another significance test is the F-test. It tests the null hypothesis:

$$H_0:R^2=0$$

<br>

This is equivalent to the following null hypothesis: 

$$H_0:\beta_1=\beta_2=\beta_3=\beta_k=0$$

<br>

The F-test statistic is calculated as follows:

\begin{equation} 
F=\frac{\frac{SS_M}{k}}{\frac{SS_R}{(n-k-1)}}=\frac{MS_M}{MS_R}
(\#eq:regSSR)
\end{equation} 

which has a F distribution with k number of predictors and n degrees of freedom. In other words, you divide the systematic ("explained") variation due to the predictor variables by the unsystematic ("unexplained") variation. 

The result of the F-test is provided in the regression output. However, you might manually compute the F-test using the ANOVA results from the model:  

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE,paged.print = FALSE}
anova(simple_regression) #anova results
f_calc <- anova(simple_regression)$'Mean Sq'[1]/anova(simple_regression)$'Mean Sq'[2] #compute F
f_calc
f_crit <- qf(.95, df1 = 1, df2 = 100) #critical value
f_crit
f_calc > f_crit #test if calculated test statistic is larger than critical value
```

#### Using the model

After fitting the model, we can use the estimated coefficients to predict sales for different values of advertising. Suppose you want to predict sales for a new product, and the company plans to spend 800 Euros on advertising. How much will it sell? You can easily compute this either by hand:

$$\hat{sales}=134.134 + 0.09612*800=211$$

<br>

... or by extracting the estimated coefficients from the model summary:

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE}
summary(simple_regression)$coefficients[1,1] + # the intercept
summary(simple_regression)$coefficients[2,1]*800 # the slope * 800
```

The predicted value of the dependent variable is 211 units, i.e., the product will (on average) sell 211 units.

**The following video summarizes how to conduct simple linear regression in R**

<div align="center">
<iframe width="560" height="315" src="https://www.youtube.com/embed/HWzvQVqGUzA" frameborder="0" allowfullscreen></iframe>
</div>

### Multiple linear regression

Multiple linear regression is a statistical technique that simultaneously tests the relationships between two or more independent variables and an interval-scaled dependent variable. The general form of the equation is given by:

\begin{equation} 
Y=(\beta_0+\beta_1*X_1+\beta_2*X_2+\beta_n*X_n)+\epsilon
(\#eq:regequ)
\end{equation} 

Again, we aim to find the linear combination of predictors that correlate maximally with the outcome variable. Note that if you change the composition of predictors, the partial regression coefficient of an independent variable will be different from that of the bivariate regression coefficient. This is because the regressors are usually correlated, and any variation in Y that was shared by X1 and X2 was attributed to X1. The interpretation of the partial regression coefficients is the expected change in Y when X is changed by one unit and all other predictors are held constant. 

Let's extend the previous example. Say, in addition to the influence of advertising, you are interested in estimating the influence of airplay on the number of album downloads. The corresponding equation would then be given by:

\begin{equation} 
Sales=\beta_0+\beta_1*adspend+\beta_2*airplay+\epsilon
(\#eq:regequadv)
\end{equation} 

The words "adspend" and "airplay" represent data that we have observed on advertising expenditures and number of radio plays, and &beta;<sub>1</sub> and &beta;<sub>2</sub> represent the unknown relationship between sales and advertising expenditures and radio airplay, respectively. The coefficients tells you by how much sales will increase for an additional Euro spent on advertising (when radio airplay is held constant) and by how much sales will increase for an additional radio play (when advertising expenditures are held constant). Thus, we can make predictions about album sales based not only on advertising spending, but also on radio airplay.

With several predictors, the partitioning of sum of squares is the same as in the bivariate model, except that the model is no longer a 2-D straight line. With two predictors, the regression line becomes a 3-D regression plane. In our example:

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE,fig.align="center", fig.cap = "Regression plane"}
#https://stackoverflow.com/questions/44322350/add-regression-plane-in-r-using-plotly
library(plotly)
library(reshape2)
n <- nrow(regression)
x1 <- regression$adspend
x2 <- regression$airplay
x3 <- regression$starpower
y <- regression$sales
df <- data.frame(y, x1, x2, x3)

### Estimation of the regression plane
mod <- lm(y ~ x1+x2)
cf.mod <- coef(mod)

### Calculate z on a grid of x-y values
x1.seq <- seq(min(x1),max(x1),length.out=25)
x2.seq <- seq(min(x2),max(x2),length.out=25)
z <- t(outer(x1.seq, x2.seq, function(x,y) cf.mod[1]+cf.mod[2]*x+cf.mod[3]*y))

#### Draw the plane with "plot_ly" and add points with "add_trace"
cols <- c("#f5cb11", "#b31d83")
cols <- cols[x3+1] 
library(plotly)
p <- plot_ly(x=~x1.seq, y=~x2.seq, z=~z,
  colors = c("#A9D0F5", "#08088A"),type="surface") %>%
  add_trace(data=df, x=x1, y=x2, z=y, mode="markers", type="scatter3d",
  marker = list(color=cols, size=3, opacity=0.8, symbol=75)) %>%
  layout(scene = list(
    aspectmode = "manual", aspectratio = list(x=1, y=1, z=1),
    xaxis = list(title = "adspend", range = c(0,2300)),
    yaxis = list(title = "airplay", range = c(0,70)),
    zaxis = list(title = "sales", range = pretty(z)[c(1,8)])))
p
```

Like in the bivariate case, the plane is fitted to the data with the aim to predict the oserved data as good as possible. The deviation of the observations from the plane represent the residuals (the error we make in predicting the observed data from the model). Note that this is conceptually the same as in the bivariate case, except that the computation is more complex (we won't go into details here). The model is fairly easy to plot using a 3-D scatterplot, because we only have two predictors. While multiple regression models that have more than two predictors are not as easy to visualize, you may apply the same principles when interpreting the model outcome:

* Total sum of squares (SS<sub>T</sub>) is still the difference between the observed data and the mean value of of Y (total variation)
* Residual sum of squares (SS<sub>R</sub>) is still the difference between the observed data and the values predicted by the model (unexplained variation)
* Model sum of squares (SS<sub>M</sub>) is still the difference between the values predicted by the model and the mean value of of Y (explained variation)
* R measures the multiple correlation between the predictors and the outcome
* R<sup>2</sup> is the amount of variation in the outcome variable explained by the model

Estimating multiple regression moodels is straightforward using the ```lm()``` function. You just need to separate the individual predictors on the right hand side of the equation using the ```+``` symbol. For example, the model:

\begin{equation} 
Sales=\beta_0+\beta_1*adspend+\beta_2*airplay+\beta_3*starpower+\epsilon
(\#eq:regequadv)
\end{equation} 

could be estimated as follows: 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
multiple_regression <- lm(sales ~ adspend + airplay + starpower, data = regression) #estimate linear model
summary(multiple_regression) #summary of results
```

The interpretation of the coefficients is as follows: 

* adspend (&beta;<sub>1</sub>): when advertising expenditures increase by 1 Euro, sales will increase by `r round(summary(multiple_regression)$coefficients[2],3)` units
* airplay (&beta;<sub>2</sub>): when radio airplay increases by 1 play per week, sales will increase by `r round(summary(multiple_regression)$coefficients[3],3)` units
* starpower (&beta;<sub>3</sub>): when the number of previous albums increases by 1, sales will increase by `r round(summary(multiple_regression)$coefficients[4],3)` units

The associated t-values and p-values are also given in the output. You can see that the p-values are smaller than 0.05 for all three coefficients. Hence, all effects are "significant". This means that if the null hypothesis was true (i.e., there was no effect between the variables and sales), the probability of observing associations of the estimated magnitudes (or larger) is very small (e.g., smaller than 0.05).     

Again, to get a better feeling for the range of values that the coefficients could take, it is helpful to compute <b>confidence intervals</b>. 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
confint(multiple_regression)
```

What does this tell you? Recall that a 95% confidence interval is defined as a range of values such that with a 95% probability, the range will contain the true unknown value of the parameter. For example, for &beta;<sub>3</sub>, the confidence interval is [`r confint(multiple_regression)[4,1]`,`r confint(multiple_regression)[4,2]`]. Thus, although we have computed a point estimate of `r round(summary(multiple_regression)$coefficients[4],3)` for the effect of starpower on sales based on our sample, the effect might actually just as well take any other value within this range, considering the sample size and the variability in our data.  

The output also tells us that `r summary(multiple_regression)$r.squared*100`% of the variation can be explained by our model. You may also visually insepct the fit of the model by plotting the predicted values agains the observed values. We can extract the predicted values using the ```predict()``` function. So let's create a new variable ```yhat```, which contains those predicted values.  

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE}
regression$yhat <- predict(simple_regression)
```

We can now use this variable to plot the predicted values against the observed values. In the following plot, the model fit would be perfect if all points would fall on the diagonal line. The larger the distance between the points and the line, the worse the model fit.

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE,fig.align="center", fig.cap = "Model fit"}
ggplot(regression,aes(yhat,sales)) +  
  geom_point(size=2,shape=1) +  #Use hollow circles
  scale_x_continuous(name="predicted values") +
  scale_y_continuous(name="observed values") +
  geom_abline(intercept = 0, slope = 1) +
  theme_bw()
```

**Partial plots**

In the context of a simple linear regression (i.e., with a single independent variable), a scatter plot of the dependent variable against the independent variable provides a good indication of the nature of the relationship. If there is more than one independent variable, however, things become more complicated. The reason is that although the scatter plot still show the relationship between the two variables, it does not take into account the effect of the other independent variables in the model. Partial regression plot show the effect of adding another variable to a model that already controls for the remaining variables in the model. In other words, it is a scatterplot of the residuals of the outcome variable and each predictor when both variables are regressed separately on the remaining predictors. As an example, consider the effect of advertising expenditures on sales. In this case, the partial plot would show the effect of adding advertising expenditures as an explanatory variable while controlling for the variation that is explained by airplay and starpower in both variables (sales and advertising). Think of it as the purified relathionship between advertising and sales that remains after controlling for other factors. The partial plots can easily be created using the ```avPlots()``` function from the ```car``` package:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Partial plots"}
library(car)
avPlots(multiple_regression)
```

**Using the model**

After fitting the model, we can use the estimated coefficients to predict sales for different values of advertising, airplay, and starpower. Suppose you would like to predict sales for a new music album with advertising expenditures of 800, airplay of 30 and starpower of 5. How much will it sell?

$$\hat{sales}=−26.61 + 0.084 * 800 + 3.367*30 + 11.08 ∗ 5= 197.74$$

<br>

... or by extracting the estimated coefficients:

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE}
summary(multiple_regression)$coefficients[1,1] + 
  summary(multiple_regression)$coefficients[2,1]*800 + 
  summary(multiple_regression)$coefficients[3,1]*30 + 
  summary(multiple_regression)$coefficients[4,1]*5
```

The predicted value of the dependent variable is 198 units, i.e., the product will sell 198 units.

**Comparing effects**

Using the output from the regression model above, it is difficult to compare the effects of the independent variables because they are all measured on different scales (Euros, radio plays, releases). Standardized regression coefficients can be used to judge the relative importance of the predictor variables. Standardization is achieved by multiplying the unstandardized coefficent by the ratio of the standard deviations of the independent and dependent variables:

\begin{equation} 
B_{k}=\beta_{k} * \frac{s_{x_k}}{s_y}
(\#eq:stdcoeff)
\end{equation}

Hence, the standardized coefficient will tell you by how many standard deviations the outcome will change as a result of a one standard deviation change in the predictor variable. Standardized coefficients can be easily computed using the ```lm.beta()``` function from the ```lm.beta``` package.

```{r message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE}
library(lm.beta)
lm.beta(multiple_regression)
```

The results show that for ```adspend``` and ```airplay```, a change by one standard deviation will result in a 0.51 standard deviation change in sales, whereas for ```starpower```, a one standard deviation change will only lead to a 0.19 standard deviation change in sales. Hence, while the effects of ```adspend``` and ```airplay``` are comparable in magnitude, the effect of ```starpower``` is less strong. 

<br>

**The following video summarizes how to conduct multiple regression in R**

<div align="center">
<iframe width="560" height="315" src="https://www.youtube.com/embed/BMKgcG9me4s" frameborder="0" allowfullscreen></iframe>
</div>

## Potential problems

Once you have built and estimated your model it is important to run diagnostics to ensure that the results are accurate. In the following section we will discuss common problems.

### Outliers

Outliers are data points that differ vastly from the trend. They can introduce bias into a model due to the fact that they alter the parameter estimates. Consider the example below. A linear regression was performed twice on the same dataset, except during the second estimation the two green points were changed to be outliers by being moved to the positions indicated in red. The solid red line is the regression line based on the unaltered dataset, while the dotted line was estimated using the altered dataset. As you can see the second regression would lead to  different conclusions than the first. Therefore it is important to identify outliers and further deal with them. 

```{r, echo = FALSE, eval = TRUE, fig.cap = "Effects of outliers", fig.align="center"}
set.seed(123)
x <- 1:30
y <- x*2.8 + rnorm(30, sd = 10)

y_altered <- y
y_altered[29] <- -50
y_altered[27] <- -47

ggplot(data = data.frame(X = x, Y = y, Y_alt = y_altered), aes(x = X, y = Y)) + 
  geom_point(aes(x = X, y = Y_alt), col = "red") + 
  geom_point() + 
  geom_point(aes(x = 27, y = Y[27]), col = "green") +
  geom_point(aes(x = 29, y = Y[29]), col = "green") +
  geom_smooth(method = "lm", se = FALSE, col = "firebrick", lwd = 0.5) + 
  geom_smooth(aes(x = X, y = Y_alt), method = "lm", se = FALSE, col = "black", lwd = 0.5, lty = 2) +
  theme_bw()


```

One quick way to visually detect outliers is by creating a scatterplot (as above) to see whether anything seems off. Another approach is to inspect the studentized residuals. If there are no outliers in your data, about 95% will be between -2 and 2, as per the assumptions of the normal distribution. Values well outisde of this range are unlikely to happen by chance and warrant further inspeciton. As a rule of thumb, observations whose studentized residuals are greater than 3 in absolute values are potential outliers.  

The studentized residuals can be obtained in R with the function ```rstudent()```. We can use this function to create a new variable that contains the studentized residuals e music sales regression from before yields the following residuals:

```{r, warning=FALSE, message=FALSE}
regression$stud_resid <- rstudent(multiple_regression)
head(regression)
```

A good way to visually inspect the studentized residuals is to plot them in a scatterplot and roughly check if most of the observations are within the -3, 3 bounds. 

```{r, warning=FALSE, message=FALSE, fig.cap="Plot of the studentized residuals", fig.align="center"}
plot(1:nrow(regression),regression$stud_resid, ylim=c(-3.3,3.3)) #create scatterplot 
abline(h=c(-3,3),col="red",lty=2) #add reference lines
```

To identify potentially influencial observations in our data set, we can apply a filter to our data:

```{r, warning = FALSE, message = FALSE}
outliers <- subset(regression,abs(stud_resid)>3)
outliers
```

After a detailed inspection of the potential outliers, you might decide to delete the affected observations from the data set or not. If an outlier has resulted from an error in data collection, then you might simply remove the observation. However, even though data may have extreme values, they might not be influential to determine a regression line. That means, the results wouldn’t be much different if we either include or exclude them from analysis. This means that the decision of whether to exclude an outlier or not is closely related to the question whether this observation is an influencial observation, as will be discussed next.  

### Influential observations

Related to the issue of outliers is that of influential observations, meaning observations that exert undue influence on the parameters. It is possible to determine whether or not the results are driven by an influencial observation by calculating how far the predicted values for your data would move if the model was fitted without this particular observation. This calculated total distance is called **Cook's distance**. To identify influencial observations, we can inspect the respective plots created from the model output. A rule of thumb to determine whether an observation should be classified as influential or not is to look for observation with a Cook's distance > 1 (although opinions vary on this). The following plot can be used to see the Cook's distance associated with each data point:

```{r, warning = FALSE, message = FALSE,fig.align="center", fig.cap = "Cook's distance"}
plot(multiple_regression,4)
```

It is easy to see that none of the Cook's distance values is close to the critical value of 1. Another useful plot to identify influential observations is plot number 5 from the output: 

```{r, warning = FALSE, message = FALSE,fig.align="center", fig.cap = "Residuals vs. Leverage"}
plot(multiple_regression,5)
```

In this plot, we look for cases outside of a dashed line, which represents **Cook’s distance**. Lines for Cook’s distance thresholds of 0.5 and 1 are included by default. In our example, this line is not even visible, since the Cook’s distance values are far away from the critical values. Generally, you would watch out for outlying values at the upper right corner or at the lower right corner of the plot. Those spots are the places where cases can be influential against a regression line. In our example, there are no influencial cases.

To see how influencial observations can impact your regression, have a look at <a href="https://en.wikipedia.org/wiki/Anscombe%27s_quartet" target="_blank">this example</a>.

### Non-linearity

An important underlying assumption for OLS is that of linearity, meaning that the relationship between the dependent and the independent variable can be reasonably approximated in linear terms. One quick way to assess whether a linear relationship can be assumed is to inspect the added variable plots that we already came across earliert:  

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Partial plots"}
library(car)
avPlots(multiple_regression)
```

In our example, it appears that linear relationships can be reasonably assumed. Please note, however, that the assumption of linearity implies two things:

* Constant marginal returns (e.g., an increase in ad-spend from 10€ to 11€ yields the same increase in sales as an increase from 100,000€ to 100,001€)
* Elasticities increase with X (e.g., advertising becomes relatively more effective; i.e., a relatively smaller change in advertising expenditure will yield the same return)

These assumptions may not be justifiable in certain contexts and you might have to transform your data (e.g., using log-transformations) in these cases, as we will see below.

### Non-constant error variance

Another important assumption of the linear model is that the error terms have a constant variance (i.e., homoscedasticity). The following plot from the model output shows the residuals (the vertical distance from an observed value to the predicted values) versus the fitted values (the predicted value from the regression model). If all the points fell exactly on the dashed grey line, it would mean that we have a perfect prediction. The residual variance (i.e., the spread of the values on the y-axis) should be similar accross the scale of the fitted values on the x-axis. 

```{r, warning = FALSE, message = FALSE,fig.align="center", fig.cap = "Residuals vs. fitted values"}
plot(multiple_regression, 1)
```

In our case, this appears to be the case. You can identify non-constant variances in the errors (i.e., heteroscedasticity) from the presence of a funnel shape in the above plot. When the assumption of constant error variances is not met, this might be due to a misspecification of your model (e.g., the relationship might not be linear). In these cases, it often helps to transform your data (e.g., using log-transformations). The red line also helps you to identify potential misspecification of your model. It is a smoothed curve that passes through the residuals and if it lies close to the gray dashed line (as in our case) it suggest a correct specification. If the line would deviate from the dashed grey line a lot (e.g., a U-shape or inverse U-shape), it would suggest that the linear model specification is not reasonable and you should try different specifications.   

If OLS is performed despite heteroscedasticity, the estimates of the coefficient will still be correct on average. However, the estimator is _inefficient_, meaning that the standard error is wrong, which will impact the significance tests (i.e., the p-values will be wrong). However, there are also robust regression methods, which you can use to estimate your model despite the presence of heteroscedasticity. 

### Non-normally distributed errors

Another assumption of OLS is that the error term is normally distributed. This can be a reasonable assumption for many scenarios, but we still need a way to check if it is actually the case. As we can not directly observe the actual error term, we have to work with the next best thing - the residuals. 

A quick way to assess whether a given sample is approximately normally distributed is by using Q-Q plots. These plot the theoretical position of the observations (under the assumption that they are normally distributed) against the actual position. The plot below is created by the model output and shows the residuals in a Q-Q plot. As you can see, most of the points roughly follow the theoretical distribution, as given by the straight line. If most of the points are close to the line, the data is approximately normally distributed.

```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", fig.cap = "Q-Q plot"}
plot(multiple_regression,2)
```

Another way to check for normal distribution of the data is to employ statistical tests that test the null hypothesis that the data is normally distributed, such as the Shapiro–Wilk test. We can extract the residuals from our model using the ```resid()``` function and apply the ```shapiro.test()``` function to it:

```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", fig.cap = "Normally distributed Q-Q plot"}
shapiro.test(resid(multiple_regression))
```

As you can see, we can not reject the H<sub>0</sub> of the normally distributed residuals, which means that we can assume the residuals to be aproximately normally distributed.

When the assumption of normally distributed errors is not met, this might again be due to a misspecification of your model, in which case it might help to transform your data (e.g., using log-transformations).

### Correlation of errors

The assumption of independent errors implies that for any two observations the residual terms should be uncorrelated. This is also known as a _lack of autocorrelation_. In theory, this could be tested with the Durbin-Watson test, which checks whether adjacent residuals are correlated. However, be aware that the test is sensitive to the order of your data. Hence, it only makes sense if there is a natural order in the data (e.g., time-series data) when the presence of dependent errors indicates autocorrelation. **Since there is no natural order in our data, we don't need to apply this test.**.

If you are confronted with data that has a natural order, you can performed the test using the command ```durbinWatsonTest()```, which takes the object that the ```lm()``` function generates as an argument. The test statistic varies between 0 and 4, with values close to 2 being desirable. As a rule of thumb values below 1 and above 3 are causes for concern. 

### Collinearity

Linear dependence of regressors, also known as multicollinearity, is when there is a strong linear relationship between the independent variables. Some correlation will always be present, but severe correlation can make proper estimation impossible. When present, it affects the model in several ways:

* Limits the size of R<sup>2</sup>: when two variables are highly correlated, the amount of unique explained variance is low; therefore the incremental change in R<sup>2</sup> by including an additional predictor is larger if the predictor is uncorrelated with the other predictors.
* Increases the standard errors of the coefficients, making them less trustworthy.
* Uncertainty about the importance of predictors: if two predictors explain similar variance in the outcome, we cannot know which of these variables is important.  

A quick way to find obvious multicollinearity is to examine the correlation matrix of the data. Any value > 0.8 - 0.9 should be cause for concern. You can, for example, create a correlation matrix using the ```rcorr()``` function from the ```Hmisc``` package.

```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
library("Hmisc")
rcorr(as.matrix(regression[,c("adspend","airplay","starpower")]))
```

The bivariate correlations can also be show in a plot:

```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", fig.cap = "Bivariate correlation plots"}
plot(regression[,c("adspend","airplay","starpower")])
```

However, this only spots bivariate multicollinearity. Variance inflation factors can be used to spot more subtle multicollinearity arising from multivariate relationships. It is calculated by regressing X<sub>i</sub> on all other X and using the resulting R<sup>2</sup> to calculate

\begin{equation} 
\begin{split}
\frac{1}{1 - R_i^2}
\end{split}
(\#eq:VIF)
\end{equation} 

VIF values of over 4 are certainly cause for concern and values over 2 should be further investigated. If the average VIF is over 1 the regression may be biased. The VIF for all variables can easily be calculated in R with the ```vif()``` function. 

```{r, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
library(car)
vif(multiple_regression)
```

As you can see the values are well below the cutoff, indicating that we don not have to worry about multicollinearity in our example.  

### Omitted Variables

If a variable that in reality influences the outcome is left out of the model ("omitted") a bias might be introduced. Specifically, the other coefficients will be biased if they are correlated with the omitted variable. Intuitively, the variables left in the model "pick up" the effect of the omitted variable to the degree that they are related. In order to understand this it helps to investigate the coefficients of the multiple regression a little bit. We have already seen that in the univariate case 

$$
\hat{\beta_1}=\frac{COV_{XY}}{s_x^2}
$$

We have [also seen that we can isolate the partial effect](#partial-plots) of a single variable. Specifically, we can calculate the coefficient of the i-th variable as

$$
\hat{\beta_{i}} = {COV(\tilde{Y_{i}}, \tilde{X_{i}}) \over V(\tilde{X_i})}
$$

where $\tilde{Y_{i}}$ is the residual from the regression of $Y$ on all variables except for the i-th and $\tilde{X_i}$ is the residual from the regression of $X_i$ on all other variables. To illustrate this let's generate some data and run a regression.

```{r }
x1 <- rnorm(100, 5, 15)
x2 <- rnorm(100, 20, 10) + 3*x1

y <- 50 + 2*x1 + 7*x2 + rnorm(100, 0, 5)
```

We have added some randomness through the ```rnorm``` command so the coefficients are not exactly as we set them but close. Clearly x1 has partial influence on y.

```{r }
mod <- lm(y~x1+x2)
summary(mod)
```

To illustrate the partial effect of x1 we run regressions for x1 on x2 and x3, as well as y on x2 and x3 and obtain the residuals $\tilde{x1}$ and $\tilde{x2}$. Notice that x1 is highly correlated with x2. 

```{r }
regX1 <- lm(x1~x2)  
summary(regX1)
tildeX1 <- residuals(regX1)
regY1 <- lm(y~x2)
tildeY1 <- residuals(regY1)
```

We can use the residuals to create the partial plots as seen above.

```{r }
# This is the same as avPlot
ggplot(data.frame(y = tildeY1, x = tildeX1), aes(x = x, y = y)) +
  geom_point(shape = 1) +
  geom_smooth(method = 'lm', se = FALSE, color = "red") +
  labs(y = "y | others", x = "x | others") + 
  theme_bw()
avPlots(mod)
```

And the regression of the residuals of y on x1 yields the same coefficient as in the original regression (minus the rounding errors). 

```{r}
# This is the same coefficient as in the original regression 
summary(lm(tildeY1~tildeX1 - 1)) # -1 since we want no intercept
cov(tildeY1, tildeX1)/var(tildeX1) 
```

The question now is what would happen if we did not include x2 in the regression? Observe that the coefficient of x1 increased substantially. That is, it picked up the effect of x2 through the covariance of x1 and x2.

```{r }
summary(lm(y~x1))
summary(mod)$coefficients[2,1] + summary(mod)$coefficients[3,1] / summary(regX1)$coefficients[2,1]
```

This estimate is inconsistent since x2 is now part of the error term and thus x1 is correlated with it. It is also quite easy to work out the bias. 

```{r }

summary(mod)$coefficients[3, 1] / summary(regX1)$coefficients[2,1] 
```



## Categorical predictors

```{r echo=F, eval=T}
regression$country <- factor(regression$country, levels = c(0:1), labels = c("local", "international")) #convert grouping variable to factor
regression$genre <- factor(regression$genre, levels = c(1:3), labels = c("rock", "pop","electronic")) #convert grouping variable to factor
```

### Two categories

Suppose, you wish to investigate the effect of the variable "country" on sales, which is a categorical variable that can only take two levels (i.e., 0 = local artist, 1 = international artist). Categorical variables with two levels are also called binary predictors. It is straightforward to include these variables in your model as "dummy" variables. Dummy variables are factor variables that can only take two values. For our "country" variable, we can create a new predictor variable that takes the form:

\begin{equation} 
x_4 =
  \begin{cases}
    1       & \quad \text{if } i \text{th artist is international}\\
    0  & \quad \text{if } i \text{th artist is local}
  \end{cases}
(\#eq:dummycoding)
\end{equation} 

This new variable is then added to our regression equation from before, so that the equation becomes 

\begin{align}
Sales =\beta_0 &+\beta_1*adspend\\
      &+\beta_2*airplay\\
      &+\beta_3*starpower\\ 
      &+\beta_4*international+\epsilon
\end{align}

where "international" represents the new dummy variable and $\beta_4$ is the coefficient associated with this variable. Estimating the model is straightforward - you just need to include the variable as an additional predictor variable. Note that the variable needs to be specified as a factor variable before including it in your model. If you haven't converted it to a factor variable before, you could also use the wrapper function ```as.factor()``` within the equation. 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
multiple_regression_bin <- lm(sales ~ adspend + airplay + starpower + country, data = regression) #estimate linear model
summary(multiple_regression_bin) #summary of results
```

You can see that we now have an additional coefficient in the regression output, which tells us the effect of the binary predictor. The dummy variable can generally be interpreted as the average difference in the dependent variable between the two groups (similar to a t-test). In this case, the coefficient tells you the difference in sales between international and local artists, and whether this difference is significant. Specifically, it means that international artists on average sell 45.67 units more than local artists, and this difference is significant (i.e., p < 0.05).  

### More than two categories

Predictors with more than two categories, like our "genre"" variable, can also be included in your model. However, in this case one dummy variable cannot represent all possible values, since there are three genres (i.e., 1 = Rock, 2 = Pop, 3 = Electronic). Thus, we need to create additional dummy variables. For example, for our "genre" variable, we create two dummy variables as follows:

\begin{equation} 
x_5 =
  \begin{cases}
    1       & \quad \text{if } i \text{th  product is from Pop genre}\\
    0  & \quad \text{if } i \text{th product is from Rock genre}
  \end{cases}
(\#eq:dummycoding1)
\end{equation} 

\begin{equation} 
x_6 =
  \begin{cases}
    1       & \quad \text{if } i \text{th  product is from Electronic genre}\\
    0  & \quad \text{if } i \text{th product is from Rock genre}
  \end{cases}
(\#eq:dummycoding2)
\end{equation} 

We would then add these variables as additional predictors in the regression equation and obtain the following model

\begin{align}
Sales =\beta_0 &+\beta_1*adspend\\
      &+\beta_2*airplay\\
      &+\beta_3*starpower\\ 
      &+\beta_4*international\\
      &+\beta_5*Pop\\
      &+\beta_6*Electronic+\epsilon
\end{align}

where "Pop" and "Rock" represent our new dummy variables, and $\beta_5$ and $\beta_6$ represent the associated regression coefficients. 

The interpretation of the coefficients is as follows: $\beta_5$ is the difference in average sales between the genres "Rock" and "Pop", while $\beta_6$ is the difference in average sales between the genres "Rock" and "Electro". Note that the level for which no dummy variable is created is also referred to as the *baseline*. In our case, "Rock" would be the baseline genre. This means that there will always be one fewer dummy variable than the number of levels.

You don't have to create the dummy variables manually as R will do this automatically when you add the variable to your equation: 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
multiple_regression <- lm(sales ~ adspend + airplay + starpower+ country + genre, data = regression) #estimate linear model
summary(multiple_regression) #summary of results
```

How can we interprete the coefficients? It is estimated based on our model that products from the "Pop" genre will on average sell 47.69 units more than products from the "Rock" genre, and that products from the "Electronic" genre will sell on average 27.62 units more than the products from the "Rock" genre. The p-value of both variables is smaller than 0.05, suggesting that there is statistical evidence for a real difference in sales between the genres.

The level of the baseline category is arbitrary. As you have seen, R simply selects the first level as the baseline. If you would like to use a different baseline category, you can use the ```relevel()``` function and set the reference category using the ```ref``` argument. The following would estimate the same model using the second category as the baseline:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
multiple_regression <- lm(sales ~ adspend + airplay + starpower+ country + relevel(genre,ref=2), data = regression) #estimate linear model
summary(multiple_regression) #summary of results
```

Note that while your choice of the baseline category impacts the coefficients and the significance level, the prediction for each group will be the same regardless of this choice.

## Extensions of the linear model

The standard linear regression model provides results that are easy to interprete and is useful to address many real-world problems. However, it makes rather restrictive assumptions that might be violated in many cases. Notably, it assumes that the relationships between the response and predictor variable is *additive* and *linear*. The additive assumption states that the effect of an inpedendent variable on the dependent variable is independent of the values of the other independent variables included in the model. The linear assumption means that the effect of a one-unit change in the independent variable on the dependent variable is the same, regardless of the values of the value of the independent variable. This is also referred to as constant *marginal returns*. For example, an increase in ad-spend from 10€ to 11€ yields the same increase in sales as an increase from 100,000€ to 100,001€. This section presents alternative model specifications if the assumptions do not hold. 

### Interaction effects

Regarding the additive assumption, it might be argued that the effect of some variables are not fully independent of the values of other variables. In our example, one could argue that the effect of advertising depends on the type of artist. For example, for local artist advertising might be more effective. We can investigate if this is the case using a grouped scatterplot:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Effect of advertising by group"}
ggplot(regression, aes(adspend, sales, colour = as.factor(country))) +
  geom_point() + 
  geom_smooth(method="lm", alpha=0.1) + 
  labs(x = "Advertising expenditures (EUR)", y = "Number of sales", colour="country") + 
  theme_bw()
```

The scatterplot indeed suggests that there is a difference in advertising effectiveness beween local and international artists. You can see this from the two different regression lines. We can incorporate this interaction effect by including an interaction term in the regression equation as follows:

\begin{align}
Sales =\beta_0 &+\beta_1*adspend\\
      &+\beta_2*airplay\\
      &+\beta_3*starpower\\ 
      &+\beta_4*international\\
      &+\beta_5*(adspend*international)\\
      &+\epsilon
\end{align}

You can see that the effect of advertising now depends on the type of artist. Hence, the additive assumption is removed. Note that if you decide to include an interaction effect, you should always include the main effects of the variables that are included in the interaction (even if the associated p-values do not suggest significant effects). It is easy to include an interaction effect in you model by adding an additional variable that has the format ```var1:var2````. In our example, this could be achieved using the following specification:  

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Effect of advertising by group"}
multiple_regression <- lm(sales ~ adspend + airplay + starpower + country + adspend:country, data = regression) #estimate linear model
summary(multiple_regression) #summary of results
```

How can we interpret the coefficient? The ```adspend``` main effect tells you the effect of advertising for the reference group that has the factor level zero. In our example, it is the advertising effect for local artist. This means that for local artists, spending an additional 1,000 Euros on advertising will result in approximately 89 additional unit sales. The interaction effect tells you by how much the effect differs for the other group (i.e., international artists) and whether this difference is significant. In our example, it means that the effect for international artists can be computed as: 0.0885 - 0.0347 = 0.0538. This means that for international artists, spending an additional 1,000 Euros on advertising will result in approximately 54 additional unit sales. Since the interaction effect is significant (p < 0.05) we can conclude that advertising is less effective for international artists.

The above example showed the interaction between a categorical variable (i.e., "country") and a continuous variable (i.e., "adspend"). However, interaction effects can be defined for different combinations of variable types. For example, you might just as well specify an interaction between two continuous variables. In our example, you might suspect that there are synergy effects between advertising expenditures and ratio airplay. It could be that advertising is more effective when an artists receives a large number of radio plays. In this case, we would specify our model as:

\begin{align}
Sales =\beta_0 &+\beta_1*adspend\\
      &+\beta_2*airplay\\
      &+\beta_3*starpower\\ 
      &+\beta_4*(adspend*airplay)\\
      &+\epsilon
\end{align}

In this case, we can interpret $\beta_4$ as the increase in the effectiveness of advertising for a one unit increase in radio airplay (or vice verca). This can be translated to R using:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
multiple_regression <- lm(sales ~ adspend + airplay + starpower + adspend:airplay, data = regression) #estimate linear model
summary(multiple_regression) #summary of results

```

However, since the p-value of the interaction is larger than 0.05, there is little statistical evidence for an interaction between the two variables.

### Non-linear relationships

In our example above, it appeared that linear relationships could be reasonably assumed. In many practical applications, however, this might not be the case. Let's review the implications of a linear specification again:

* Constant marginal returns (e.g., an increase in ad-spend from 10€ to 11€ yields the same increase in sales as an increase from 100,000€ to 100,001€)
* Elasticities increase with X (e.g., advertising becomes relatively more effective; i.e., a relatively smaller change in advertising expenditure will yield the same return)

In many marketing contexts, these might not be reasonable assumptions. Consider the case of advertising. It is unlikely that the return on advertising will not depend on the level of advertising expenditures. It is rather likely that saturation occurs at some level, meaning that the return from an additional Euro spend on advertising is decreasing with the level of advertising expenditures (i.e., decreasing marginal returns). In other words, at some point the advertising campaign has achieved a certail level of penetration and an additional Euro spend on advertising won't yield the same return as in the beginning.

Let's use an example data set, containing the advertising expenditures of a company and the sales (in thousand units).

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
non_linear_reg <- read.table("https://raw.githubusercontent.com/IMSMWU/Teaching/master/MRDA2017/non_linear.dat", 
                          sep = "\t", 
                          header = TRUE) #read in data
head(non_linear_reg)
```

Now we inspect if a linear specification is appropriate by looking at the scatterplot:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Non-linear relationship"}
ggplot(data = non_linear_reg, aes(x = advertising, y = sales)) +
  geom_point(shape=1) + 
  geom_smooth(method = "lm", fill = "blue", alpha=0.1) + 
  theme_bw()
```

It appears that a linear model might **not** represent the data well. It rather appears that the effect of an additional Euro spend on advertising is decreasing with increasing levels of advertising expenditures. Thus, we have decreasing marginal returns. We could put this to a test and estimate a linear model:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
linear_reg <- lm(sales ~ advertising, data = non_linear_reg)
summary(linear_reg)
```

Advertising appears to be positively related to sales with an additional Euro that is spent on advertising resulting in 0.0005 additional sales. The R<sup>2</sup> statistic suggests that approximately 51% of the total variation can be explained by the model

To test if the linear specification is appropriate, let's inspect some of the plots that are generated by R. We start by inspecting the residuals plot. 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Residuals vs. Fitted"}
plot(linear_reg,1)
```

The plot suggests that the assumption of homescedasticity is violated (i.e., the spread of values on the y-axis is different for different levels of the fitted values). In addition, the red line deviates from the dashed grey line, suggesting that the relationship might not be linear. Finally, the Q-Q plot of the residuals suggests that the residuals are not normally distributed. 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Q-Q plot"}
plot(linear_reg,2)
```

To sum up, a linear specification might not be the best model for this data set. 

In this case, a multiplicative model might be a better representation of the data. The multiplicative model has the following formal representation: 

\begin{equation} 
Y =\beta_0 *X_1^{\beta_1}*X_2^{\beta_2}*...*X_J^{\beta_J}*\epsilon
(\#eq:multiplicative)
\end{equation} 

This functional form can be linearized by taking the logarithm of both sides of the equation:

\begin{equation} 
log(Y) =log(\beta_0) + \beta_1*log(X_1) + \beta_2*log(X_2) + ...+ \beta_J*log(X_J) + log(\epsilon)
(\#eq:multiplicativetransformed)
\end{equation}

This means that taking logarithms of both sides of the equation makes linear estimation possible. Let's test how the scatterplot would look like if we use the logarithm of our variables using the ```log()``` function instead of the original values.  

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Linearized effect"}
ggplot(data = non_linear_reg, aes(x = log(advertising), y = log(sales))) + 
  geom_point(shape=1) + 
  geom_smooth(method = "lm", fill = "blue", alpha=0.1) +
  theme_bw()
```

It appears that now, with the log-transformed variables, a linear specification is a much better representation of the data. Hence, we can log-transform our variables and estimate the following equation: 

\begin{equation} 
log(sales) = log(\beta_0) + \beta_1*log(advertising) + log(\epsilon)
(\#eq:multiplicativetransformed1)
\end{equation}

This can be easily implemented in R by transforming the variables using the ```log()``` function:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE}
log_reg <- lm(log(sales) ~ log(advertising), data = non_linear_reg)
summary(log_reg)
```

Note that this specification implies decreasing marginal returns (i.e., the returns of advertising are decreasing with the level of advertising), which appear to be more consistent with the data. The specification is also consistent with proportional changes in advertising being associated with proportional changes in sales (i.e., advertising does not become more effective with increasing levels). This has important implications on the interpretation of the coefficients. In our example, you would interpret the coefficient as follows: **A 1% increase in advertising leads to a 0.3% increase in sales**. Hence, the interpretation is in proportional terms and no longer in units. This means that the coefficients in a log-log model can be directly interpreted as elasticities, which also makes communication easier. We can genreally also inspect the R<sup>2</sup> statistic to see that the model fit has inreased compared to the linear specification (i.e., R<sup>2</sup> has increased to 0.681 from 0.509). However, please note that the variables are now measured on a different scale, which means that the model fit in theory is not directly comparable. Also, we could use the residuals plot to confirm that the revised specification is more appropriate:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center", fig.cap = c("Residuals plot","Q-Q plot")}
plot(log_reg,1)
plot(log_reg,2)
```

Finally, we can plot the predicted values against the observed values to see that the results from the log-log model (red) provide a better prediction than the results from the linear model (blue). 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE,fig.align="center", fig.cap = "Comparison if model fit"}
non_linear_reg$pred_lin_reg <- predict(linear_reg)
non_linear_reg$pred_log_reg <- predict(log_reg)
ggplot(data = non_linear_reg) +
  geom_point(aes(x = advertising, y = sales),shape=1) + 
  geom_line(data = non_linear_reg,aes(x=advertising,y=pred_lin_reg),color="blue", size=1.05) + 
  geom_line(data = non_linear_reg,aes(x=advertising,y=exp(pred_log_reg)),color="red", size=1.05) + theme_bw()
```

