---
title: "07-introStatTests"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction to Statistical Inference

This chapter will provide you with a basic intuition on statistical inference. As marketing researchers we are usually faced with "imperfect" data in the sense that we cannot collect **all** the data we would like. Imagine you are interested in the average amount of time WU students spend listening to music every month. Ideally we could force all WU students to fill out our survey. Realistically we will only be able to observe a small fraction of students (maybe 500 out of the $25.000+$). With the data from this small fraction at hand, we want to make an inference about the true average listening time of all WU students. We are going to start with the assumption that we know everything. That is, we first assume that we know all WU students' listening times and analyze the distribution of the listening time in the entire population. Subsequently, we are going to look at the uncertainty that is introduced by only knowing only some of the students' listening times (i.e., a sample from the population) and how that influences our analysis.

## If we knew it all 

Assume there are $25,000$ students at WU and every single one has kindly provided us with the hours they listened to music in the past month. In this case we know the true mean ($49.93$ hours) and the true standard deviation (SD = $10.02$) and thus we can easily summarize the entire distribution. Since the data follows a normal distribution, roughly 95% of the values lie within 2 standard deviations from the mean, as the following plot shows:

```{r, message = FALSE, warning=FALSE}
library(tidyverse)
set.seed(321)
hours <- rnorm(25000, 50, 10)

ggplot(data.frame(hours)) +
  geom_histogram(aes(hours), bins = 50, fill = 'white', color = 'black') +
  ggtitle("Histogram of listening time") +
  labs(y = 'Number of students', x = 'Hours') +
  theme_bw() +
  geom_vline(xintercept = mean(hours), size = 1) +
  geom_vline(xintercept = mean(hours)+2*sd(hours), colour = "red", size = 1) +
  geom_vline(xintercept = mean(hours)-2*sd(hours), colour = "red", size = 1) +
  annotate("text", x = mean(hours) + 43, y = 1750, 
           label = paste("Mean:", round(mean(hours), 2))) +
  annotate("text", x = mean(hours) + 44, y = 1650,
           label = paste("SD:", round(sd(hours), 2))) +
  geom_segment(aes(x = mean(hours), y = 1100, yend = 1100, xend = (mean(hours) - 2*sd(hours))), lineend = "butt", linejoin = "round",
     size = 0.5, arrow = arrow(length = unit(0.2, "inches"))) +
  geom_segment(aes(x = mean(hours), y = 1100, yend = 1100, xend = (mean(hours) + 2*sd(hours))), lineend = "butt", linejoin = "round",
     size = 0.5, arrow = arrow(length = unit(0.2, "inches"))) +
  annotate("text", x = mean(hours) + 28, y = 1100, label = "Mean + 2 * SD" )+
  annotate("text", x = mean(hours) -28, y = 1100, label = "Mean - 2 * SD" )
```

In this case, we refere to all WU students as **the population**. In general, the population is the entire group we are interested in. This group does not necessarily consist of people, but could also be companies, stores, animals, etc.. The parameters of the distribution of population values (in hour case: "hours") are called population parameters. As already mentioned, we do not usually know population parameters but use inferencial statistics to infer them based on our sample from the population. Here, we will use the following notation to refer to either the population or the sample: 

Variable	 | Population	 | Sample
------------------------- | ------------------------- | -------------------------
Size  | N  | n  
Mean  | $\mu = {1 \over N}\sum_{i=1}^N x_i$  | $\overline{x} = {1 \over n}\sum_{i=1}^n x_i$
Variance  | $\sigma^2 = {1 \over N}\sum_{i=1}^N (x_i-\mu)^2$  | $s^2 = {1 \over n-1}\sum_{i=1}^n (x_i-\overline{x})^2$
Standard deviation  | $\sigma = \sqrt{\sigma^2}$  | $s = \sqrt{s^2}$

### Sampling from a known population

In the first step towards a realistic reasearch setting, let us take samples from this distribution and calculate the mean in each sample to get an idea of how much uncertainty we introduce by only knowing a part of the population. 

Let's first randomly sample 100 students without replacement (that is once a student has been drawn she or he is removed from the pool and cannot be drawn again) and calculate the mean. We can simply sample the row numbers of students and then subset the ```hours``` vector with the sampled rownumbers. The ```sample()``` function wil be used to draw a sample of size 100 from the population of 25,000 students, and one student can only be drawn once (i.e., ```replace = FALSE```).

```{r}
student_sample <- sample(1:25000, size = 100, replace = FALSE)
(m1 <- mean(hours[student_sample]))
```

Observe that in this first draw the mean is quite close to the actual mean. It seems like the sample mean is a decent estimate of the population mean. However, we could just be lucky this time and the next sample turns out to have a completely different mean. So in order to make sure that the first draw is not just pure luck and the sample mean is in fact a good estimate for the population mean let's take **many** (e.g. $20,000$) different random samples and calculate their means. This will show us a range within which the sample mean of any sample we take is likely going to be. We are going to store the means of all the samples in a matrix and then plot a histogram of the means to observe the likely values. The sample size is $100$ for each sample that we take. 

```{r}
set.seed(12345)
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

meansdf <- data.frame('true' = mean(hours), 'sample' = mean(means))
meansdf <- gather(meansdf)
ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.title = element_blank()) +
  ggtitle('Distribution of sample means')
```

As you can see, on average the sample mean ("mean of sample means") is extremely close to the population mean, despite only sampling $100$ people at a time. This distribution of sample means is also referred to as **sampling distribution** of the sample mean. However, there is some uncertainty, as visualized by the histogram. Let's look at the means from the first $5$ samples, as well as the minimum and the maximum of the sample means:

```{r}
head(means, 5)
min(means)
max(means)
```

As you can see, the means are slightly different for the different samples and range from `r round(min(means),2)` to `r round(max(means),2)`. The following plot shows the distribution of listening times for four different samples from the population.  

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.height = 6, fig.width = 8}
#student_sample <- sample(1:25000, size = 100, replace = FALSE)
#means <- hours[student_sample]

set.seed(8830)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot1 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

set.seed(6789)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot2 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

set.seed(3904)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot3 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

set.seed(3333)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot4 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

p <- plot_grid(plot1, plot2, plot3, plot4, ncol = 2,
           labels = c("A", "B","C","D"))
title <- ggdraw() + draw_label('Distribution of listening times in four different samples', fontface='bold')
p <- plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
print(p)

```

Again, it becomes clear that the sample means are slightly different. This is referred to as **sampling variation** and it is completely fine to get a slightly different mean every time we take a sample. We just need to find a way of expressing the uncertainty associated with the fact that we only have data from one sample because in a realistic setting, you are most likely only going to have access to a single sample. 

Due to the variation in the sample means shown in our simulation, it is never possible to say exactly what the population mean is. However, even with a single sample we can infer a range of values within which the population mean is likely contained. In order to do so, notice that the sample means are approximately normally distributed. Their mean is roughly equal to the population mean, but in order to make statements about the expected range of values, we would need to know the standard deviation of the sampling distribution. Intuitively, the larger the size of the sample we take every time, the less the standard deviation is going to be. Think of the two extremes: sample size $1$ and sample size $25,000$. With a single person in the sample we do not gain a lot of information and our estimate is very uncertain, which is expressed through a larger standard deviation. Looking at the first histogram showing the number of students for each of the listening times, clearly we would get values below $25$ or above $75$ for some samples. This is way farther away from the population mean than the minimum and the maximum of our $100$ person samples. On the other hand, if we sample every student we get the population mean every time and thus we do not have any uncertainty (assuming the population does not change). Even if we only sample, say $24,000$ people every time we gain a lot of information about the population and the sample means would not be very different from each other since only up to $1,000$ people are potentially different in any given sample.

The following plots show the relationship between the sample size and the range of sample means resulting from the repeated sampling process. Notice that the more students are contained in the individual samples, the less uncertainty there is when estimating the population mean from a sample (i.e., the possible values are more closely centered around the mean). So when the sample size is small, the sample mean can expected to be very different the next time we take a sample. When the sample size is large, we can expect the sample means to be more similar.

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.height = 6, fig.width = 8}
library(cowplot)
library(gridExtra)
library(grid)
library(egg)
#install.packages("egg")
set.seed(12345)

sample_size = 10
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = sample_size, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}
meansdf <- data.frame('true' = mean(hours), 'sample' = mean(means))
meansdf <- gather(meansdf)
plot1 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.position = "none") +  ggtitle('Sample size = 10')

sample_size = 100
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = sample_size, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}
meansdf <- data.frame('true' = mean(hours), 'sample' = mean(means))
meansdf <- gather(meansdf)
plot2 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.position = "none") +  ggtitle('Sample size = 100')

sample_size = 1000
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = sample_size, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}
meansdf <- data.frame('true' = mean(hours), 'sample' = mean(means))
meansdf <- gather(meansdf)
plot3 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.position = "none") +  ggtitle('Sample size = 1,000')

sample_size = 10000
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = sample_size, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}
meansdf <- data.frame('true' = mean(hours), 'sample' = mean(means))
meansdf <- gather(meansdf)
plot4 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.position = "none") +
  ggtitle('Sample size = 10,000')

p <- plot_grid(plot1, plot2, plot3, plot4, ncol = 2,
           labels = c("A", "B","C","D"))
title <- ggdraw() + draw_label('Distribution of sample means', fontface='bold')
p <- plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
print(p)
# now add the title
#title <- ggdraw() + draw_label("", fontface='bold')
#plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins

```

A second factor determining the standard deviation of the distribution of sample means is the standard deviation associated with the population parameter. Again, looking at the extremes illustrates this well. If all WU students listened to music for approximately the same amount of time, the samples would not differ much from each other. In other words, if the standard deviation in the population is lower, we expect the standard deviation of the sample means to be lower as well. This is illustrated by the following plots. In the first plot, we assume a much smaller population standard deviation (i.g., SD = 1 instead of SD = 10). Notice how the smaller (larger) the population standard deviation, the less (more) uncertainty there is when estimating the population mean from a sample (i.e., the possible values are more closely centered around the mean). So when the population SD is large, the sample mean can expected to be very different the next time we take a sample. When the population SD is small, we can expect the sample means to be more similar.   

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.height = 3, fig.width = 8}
library(cowplot)
library(gridExtra)
library(grid)
library(egg)
library(truncnorm)
#install.packages("egg")
set.seed(12345)

hours1 <- rnorm(25000, 50, 1)
sample_size = 100
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = sample_size, replace = FALSE)
  means[i,] <- mean(hours1[student_sample])
}
meansdf <- data.frame('true' = mean(hours1), 'sample' = mean(means))
meansdf <- gather(meansdf)
plot1 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.position = "none") +  ggtitle('Population SD = 1')


hours2 <- rnorm(25000, 50, 10)
sample_size = 100
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = sample_size, replace = FALSE)
  means[i,] <- mean(hours2[student_sample])
}
meansdf <- data.frame('true' = mean(hours2), 'sample' = mean(means))
meansdf <- gather(meansdf)
plot2 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() +
  geom_vline(data = meansdf, aes(xintercept = value, color = key, linetype = key), size=1) +
  scale_color_discrete(labels = c("Mean of sample means", "Population mean")) +
  scale_linetype_discrete(labels = c("Mean of sample means", "Population mean")) +
  theme(legend.position = "none") +  ggtitle('Population SD = 10')

p <- plot_grid(plot1, plot2, ncol = 2, 
           labels = c("A", "B"))
title <- ggdraw() + draw_label('Distribution of sample means', fontface='bold')
p <- plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
print(p)
# now add the title
#title <- ggdraw() + draw_label("", fontface='bold')
#plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins

```

Following the notion that the population standard deviation and the sample size influences the distribution of (hypothetical) sample means, as represented by the sampling distribution, the formal representaion standard deviation of the sample means is

$$
\sigma_{\bar x} = {\sigma \over \sqrt{n}}
$$

where $\sigma$ is the population SD and $n$ is the sample size. $\sigma_{\overline{x}}$ is referred to as the **Standard Error (SE)** of the mean and it expresses the variation we would naturally expect to find, given the number of observations and the population SD. Therefore, an increase in the population SD also increases the SD of the sample mean and an increase in the number of observations per sample decreases the SD of the sample mean. 

## The Central Limit Theorem

The attentive reader might have noticed that the population above was generated using a normal density. It would be very restrictive if we could only analyze populations whose values are normally distributed. Furthermore, we are unable in reality to check whether the population values are normally distributed since we do not know the entire population. However, it turns out that the results generalize to any distribution for which the mean exists. This is described by the **Central Limit Theorem**. 

The central limit theorem states that if **(1)** the population distribution has a mean, and **(2)** we take a large enough sample, then the sampling distribution of the sample mean is approximatelly normally distributed. What exactly "large enough" means depends on the setting, but the interactive element below illustrates how the sample size influences how accurately we can estimate the population parameters from the sample statistics. 

To illustrate this, let's repeat the analysis above with a population from a gamma distribution. In the previous example we assumed a normal distribution so it was more likely for a given student to spend around 50 hours per week listening to music. This example depicts the case in which most students have a similar music listening time while some music ethusiasts have a much higher listening time. Here is a histogram of the listening times in the population:

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center", fig.height = 3, fig.width = 5}
set.seed(321)
hours <- rgamma(25000, shape = 2, scale = 10)
ggplot(data.frame(hours)) +
  geom_histogram(aes(x = hours), bins = 30, fill='white', color='black') +
    geom_vline(xintercept = mean(hours), size = 1)  +  theme_bw() +
  ggtitle('Histogram of listening times')
```

The following plot depicts the historam of listening times of one random sample of 100 students from the population: 

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.height = 6, fig.width = 8}
#student_sample <- sample(1:25000, size = 100, replace = FALSE)
#means <- hours[student_sample]

set.seed(8830)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot1 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

set.seed(6789)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot2 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

set.seed(3904)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot3 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

set.seed(3333)
student_sample <- sample(1:25000, size = 100, replace = FALSE)
means1 <- hours[student_sample]
plot4 <- ggplot(data.frame(means1)) +
  geom_histogram(aes(x = means1), bins = 30, fill='white', color='black') +
  theme_bw() + xlab("Hours") +
  geom_vline(aes(xintercept = mean(means1)), size=1) +
  ggtitle(paste('Mean: ',round(mean(means1),2)))

p <- plot_grid(plot1, plot2, plot3, plot4, ncol = 2,
           labels = c("A", "B","C","D"))
title <- ggdraw() + draw_label('Distribution of listening times in four different samples', fontface='bold')
p <- plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
print(p)

```

As you can see, the distribution of listening times no longer follows a normal distribution as a result of the fact that we now assume a gamma distribution for the population with a positive skew (i.e., lower values more likely, higher values less likely). For each individual sample that we may take in real life, this is the **natural variation** that we observe in our data.

But what about the **variation bewtween study results**, i.e., if we would (hypothetically) repeat the study many times? Let's see what happens to the distribution of sample means if we were to take an increasing number of samples, each drawn from the same gamma population:

```{r message=FALSE, warning=FALSE, echo=FALSE, eval=TRUE, fig.align="center", fig.height = 6, fig.width = 8}
library(cowplot)
library(gridExtra)
library(grid)
library(egg)
#install.packages("egg")
set.seed(12345)

hours <- rgamma(25000, shape = 2, scale = 10)

samples <- 10
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

plot1 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() + xlim(14, 26) + 
  theme(legend.title = element_blank()) +
  ggtitle('10 samples') 

samples <- 100
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

plot2 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() + xlim(14, 26) +
  theme(legend.title = element_blank()) +
  ggtitle('100 samples') 

samples <- 1000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

plot3 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() + xlim(14, 26) +
  theme(legend.title = element_blank()) +
  ggtitle('100 samples') 

samples <- 10000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

plot4 <- ggplot(data.frame(means)) +
  geom_histogram(aes(x = means), bins = 30, fill='white', color='black') +
  theme_bw() + xlim(14, 26) +
  theme(legend.title = element_blank()) +
  ggtitle('1000 samples') 

p <- plot_grid(plot1, plot2, plot3, plot4, ncol = 2,
           labels = c("A", "B","C","D"))
title <- ggdraw() + draw_label('Distribution of sample means from gamma population', fontface='bold')
p <- plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)) # rel_heights values control title margins
print(p)

```

As you can see, the more (hypothetical) samples we take, the more the sampling distribution approaches a normal distribution. Thus, when we take a large number of samples (e.g. $20,000$), the sampling distribution will be normally distributed, as the following plot shows where the red line represents the theoretical normal density function: 

```{r message=FALSE, warning=FALSE, echo=TRUE, eval=TRUE, fig.align="center", fig.height = 3, fig.width = 8}
set.seed(321)
hours <- rgamma(25000, shape = 2, scale = 10)
samples <- 20000
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

ggplot(data.frame(means)) +
  geom_histogram(aes(x = means, y = ..density..), bins = 30, fill='white', color='black') +
  stat_function(xlim = c(14, 26), fun = dnorm, n = length(means), 
                args = list(mean = mean(hours), sd = sd(hours)/ sqrt(100)), 
                aes(color = 'Theoretical density')) +
  theme_bw() +
  theme(legend.title = element_blank()) +
  ggtitle('Distribution of sample means with gamma population') 
```

To summarize, the sampling distribution (i.e., the distribution of hypothetical study means) follows a normal distribution, even though the data from each individual study are non-normal. This is an important feature, since many parametric statistical tests assume a normally distributed sampling distribution. As we have seen, this does **not** mean that your data needs to follow a normal distribution. 

## Using what we actually know

So far we have assumed to know the population standard deviation. This an unrealistic assumption since we do not know the entire population. The best guess for the population standard deviation we have is the sample standard deviation, denoted $s$. Thus, the **Standard Error (SE)** of the mean is usually estimated from the sample standard deviation: 

$$
\sigma_{\bar x} \approx {s \over \sqrt{n}}
$$

Note that $s$ itself is a sample estimate of the population parameter $\sigma$. This additional estimation introduces new uncertainty. You can see in the interactive element below that the sample SD, on average, provides a good estimate of the population SD. That is the distribution of sample SDs that we get by drawing many samples is centered around the population value. Again, the larger the sample, the closer any given sample SD is going to be to the population parameter and we introduce less uncertainty. One conclusion is that your sample needs to be large enough to provide a reliable estimate of the population parameters. What exactly "large enough" means depends on the setting, but the interactive element illustrates how the remaining values change as a function of the sample size.

We will not go into detail about the importance of random samples but basically the correctness of your estimate depends crucially on having a sample at hand that actually represents the population. Unfortunatelly we will usually not notice if the sample is non-random. Our statistics are still a good approximation of "a" population parameter, namely the one for the population that we actually sampled but not the one we are interested in. To illustrate this uncheck the "Random Sample" box below. The new sample will be only from the top $50\%$ music listeners (but this generalizes to different types of non-random samples).

<iframe src="https://learn.wu.ac.at/shiny/imsm/clt/" style="border: none; width: 800px; height: 1265px"></iframe>

### Confidence Intervals for the Sample Mean

So far, we have seen that we measure statistics from samples (e.g., the sample mean $\bar x$) in order to determine parameters of populations (e.g., the population mean $\mu$). As we have seen, the average value from a sample only provides an estimate of what the real population parameter is. The next time you collect a sample of the same size, you could get a different average (i.e., sampling variation). Hence, coming up with an exact estimate (i.e., a **point estimate**) for a particular population parameter is difficult due to the uncertainty that arises because we only have access to one sample. That is why it is often informative to construct a range around that statistic (i.e., an **interval estimate**) that likely contains the population parameter with a certain level of confidence. That is, we construct an interval such that for a large share (say 95%) of the sample means we could potentially get, the population mean is within that interval. To construct this interval, we can use our knowledge about the sample mean, the estimate of the population, and combine it with our knowledge about the distribution of interest. In the case of the normal distribution, approximately $95\%$ of the density is within $\pm 1.96$ standard deviations from the mean. Since we have estimates for the mean ($\bar x$) and population SD ($\sigma_{\bar x}$), we can calculate the lower and upper boudaries of our confidence interval as: 

```{r, fig.height = 15, fig.width=10}
pnorm(1.96)

```

$$
CI_{lower} = \bar x_1 - 1.96 * \sigma_{\bar x} \\
CI_{upper} = \bar x_1 + 1.96 * \sigma_{\bar x}
$$ 

such that in $95\%$ of the cases, the population mean is going to be within this interval. This is illustrated in the plot below that shows the mean of the first 100 samples and their confidence intervals.

```{r, fig.height = 15, fig.width=10}
set.seed(12345)
samples <- 100
hours <- rnorm(25000, 50, 10)
means <- matrix(NA, nrow = samples)
for (i in 1:samples){
  student_sample <- sample(1:25000, size = 100, replace = FALSE)
  means[i,] <- mean(hours[student_sample])
}

means_sd <- data.frame(means, lower =  means - 2 * (sd(hours)/sqrt(100)), upper = means + 2 * (sd(hours)/sqrt(100)), y = 1:100)
means_sd$diff <- factor(ifelse(means_sd$lower > mean(hours) | means_sd$upper < mean(hours), 'No', 'Yes'))

ggplot(means_sd, aes(y = y)) +
  scale_y_continuous(breaks = seq(1, 100, by = 1), expand = c(0.005,0.005)) +
  geom_point(aes(x = means, color = diff)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper, color = diff)) +
  geom_vline(xintercept = mean(hours)) +
  scale_color_manual(values = c("red", "black")) +
  guides(color=guide_legend(title="True mean in CI")) +
  theme_bw()

```

Note that this does **not** mean that for a specific sample there is a $95\%$ chance that the population mean lies within its confidence interval. The statment depends on the large number of samples we do not actually draw in a real setting. You can view the set of all possible confidence intervals similarly to the sides of a coin or a die. If we throw a coin many times, we are going to observe head roughly half of the times. This does not, however, exclude the possiblity of observing tails for the first 10 throws. Similarly, any specific confidence interval might or might not include the population mean but if we take many samples on average $95\%$ of the confidence intervals are going to include the population mean.

## Summary

When conducting research, we are (almost) always faced with the problem of not having access to the entire group we are interested in. Therefore we use sample properties that we have derived in this chapter to do statistical inference based on a single sample. We use statistics of the sample as well as the sample size to calculate the confidence interval of our choice (e.g. $95\%$). In practice most of this is done for us, so you won't need to do this by hand. With a sample at hand we need to choose the appropriate test and a null hypothesis. How this is done will be discussed in the next chapter. 
