---
output:
  html_document:
    toc: yes
  html_notebook: default
  pdf_document:
    toc: yes
---

```{r eval=TRUE, echo=F, message=FALSE, warning=FALSE}
library(knitr)
options(digits = 7, scipen = 999)
opts_chunk$set(tidy.opts=list(width.cutoff=75),tidy=FALSE, rownames.print = FALSE, rows.print = 10, echo = TRUE, warning = FALSE, message = FALSE)
```

```{r eval=TRUE, echo=F, message=FALSE, warning=FALSE}
unloadNamespace("latex2exp") # unloads latex2exp to fix the specific error, that the chart in 6.6.2 didn't display TeX output properly
```

## Logistic regression

**The following video summarizes how to visualize log-transformed regressions in R**

<div align="center">
<iframe width="560" height="315" src="https://www.youtube.com/embed/phpl7vEJUQY" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
</div>

### Motivation and intuition

In the last section we saw how to predict continuous outcomes (sales, height, etc.) via linear regression models. Another interesting case is that of binary outcomes, i.e. when the variable we want to model can only take two values (yes or no, group 1 or group 2, dead or alive, etc.). To this end we would like to estimate how our predictor variables change the probability of a value being 0 or 1. In this case we can technically still use a linear model (e.g. OLS). However, its predictions will most likely not be particularly useful. A more useful method is the logistic regression. In particular we are going to have a look at the logit model. In the following dataset we are trying to predict whether a song will be a top-10 hit on a popular music streaming platform. In a first step we are going to use only the danceability index as a predictor. Later we are going to add more independent variables. 

```{r, echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.cap="Simulated binary outcome data"}
library(ggplot2)
library(gridExtra)

chart_data <- read.delim2("https://raw.githubusercontent.com/IMSMWU/MRDA2018/master/data/chart_data_logistic.dat",header=T, sep = "\t",stringsAsFactors = F, dec = ".")
#Create a new dummy variable "top10", which is 1 if a song made it to the top10 and 0 else:
chart_data$top10 <- ifelse(chart_data$rank<11,1,0)

# Inspect data
head(chart_data)
str(chart_data)
```

Below are two attempts to model the data. The left assumes a linear probability model (calculated with the same methods that we used in the last chapter), while the right model is a __logistic regression model__. As you can see, the linear probability model produces probabilities that are above 1 and below 0, which are not valid probabilities, while the logistic model stays between 0 and 1. Notice that songs with a higher danceability index (on the right of the x-axis) seem to cluster more at $1$ and those with a lower more at $0$ so we expect a positive influence of danceability on the probability of a song to become a top-10 hit. 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.cap="The same binary data explained by two models; A linear probability model (on the left) and a logistic regression model (on the right)"}
#Scatterplot showing the association between two variables using a linear model
plt.lin <- ggplot(chart_data,aes(danceability,top10)) +  
  geom_point(shape=1) +
  geom_smooth(method = "lm") +
  theme_bw()

#Scatterplot showing the association between two variables using a glm
plt.logit <- ggplot(chart_data,aes(danceability,top10)) +  
  geom_point(shape=1) +
  geom_smooth(method = "glm", 
              method.args = list(family = "binomial"), 
              se = FALSE) +
  theme_bw()
grid.arrange(plt.lin, plt.logit, ncol = 2)
```

A key insight at this point is that the connection between $\mathbf{X}$ and $Y$ is __non-linear__ in the logistic regression model. As we can see in the plot, the probability of success is most strongly affected by danceability around values of $0.5$, while higher and lower values have a smaller marginal effect. This obviously also has consequences for the interpretation of the coefficients later on.  

### Technical details of the model

As the name suggests, the logistic function is an important component of the logistic regression model. It has the following form:

$$
f(\mathbf{X}) = \frac{1}{1 + e^{-\mathbf{X}}}
$$
This function transforms all real numbers into the range between 0 and 1. We need this to model probabilities, as probabilities can only be between 0 and 1. 

```{r, echo = FALSE}
library(latex2exp)
x <- seq(-10, 10, length.out = 1000)
fx <- 1/(1+exp(-x))
df <- data.frame(x = x, fx = fx)
ggplot(df, aes(x = x, y = fx)) + 
  geom_line()+
  labs(y = TeX("$\\frac{1}{1+e^{-\\mathbf{X}}}$"), x = TeX("$\\mathbf{X}$"))+
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) + 
  theme_bw()
```



The logistic function on its own is not very useful yet, as we want to be able to determine how predictors influence the probability of a value to be equal to 1. To this end we replace the $\mathbf{X}$ in the function above with our familiar linear specification, i.e.

$$
\mathbf{X} = \beta_0 + \beta_1 * x_{1,i} + \beta_2 * x_{2,i} + ... +\beta_m * x_{m,i}\\
f(\mathbf{X}) = P(y_i = 1) = \frac{1}{1 + e^{-(\beta_0 + \beta_1 * x_{1,i} + \beta_2 * x_{2,i} + ... +\beta_m * x_{m,i})}}
$$

In our case we only have $\beta_0$ and $\beta_1$, the coefficient associated with danceability. 

In general we now have a mathematical relationship between our predictor variables $(x_1, ..., x_m)$ and the probability of $y_i$ being equal to one. The last step is to estimate the parameters of this model $(\beta_0, \beta_1, ..., \beta_m)$ to determine the magnitude of the effects.  

### Estimation in R

We are now going to show how to perform logistic regression in R. Instead of ```lm()``` we now use ```glm(Y~X, family=binomial(link = 'logit'))``` to use the logit model. We can still use the ```summary()``` command to inspect the output of the model. 

```{r}
#Run the glm
logit_model <- glm(top10 ~ danceability,family=binomial(link='logit'),data=chart_data)
#Inspect model summary
summary(logit_model )
```

Noticeably this output does not include an $R^2$ value to asses model fit. Multiple "Pseudo $R^2$s", similar to the one used in OLS, have been developed. There are packages that return the $R^2$ given a logit model (see ```rcompanion``` or ```pscl```). The calculation by hand is also fairly simple. We define the function ```logisticPseudoR2s()``` that takes a logit model as an input and returns three popular pseudo $R^2$ values.

```{r}
logisticPseudoR2s <- function(LogModel) {
  dev <- LogModel$deviance 
  nullDev <- LogModel$null.deviance 
  modelN <- length(LogModel$fitted.values)
  R.l <-  1 -  dev / nullDev
  R.cs <- 1- exp ( -(nullDev - dev) / modelN)
  R.n <- R.cs / ( 1 - ( exp (-(nullDev / modelN))))
  cat("Pseudo R^2 for logistic regression\n")
  cat("Hosmer and Lemeshow R^2  ", round(R.l, 3), "\n")
  cat("Cox and Snell R^2        ", round(R.cs, 3), "\n")
  cat("Nagelkerke R^2           ", round(R.n, 3),    "\n")
}
#Inspect Pseudo R2s
logisticPseudoR2s(logit_model )
```

The coefficients of the model give the change in the [log odds](https://en.wikipedia.org/wiki/Odds#Statistical_usage) of the dependent variable due to a unit change in the regressor. This makes the exact interpretation of the coefficients difficult, but we can still interpret the signs and the p-values which will tell us if a variable has a significant positive or negative impact on the probability of the dependent variable being $1$. In order to get the odds ratios we can simply take the exponent of the coefficients. 

```{r}
exp(coef(logit_model ))
```

Notice that the coefficient is extremely large. That is (partly) due to the fact that the danceability variable is constrained to values between $0$ and $1$ and the coefficients are for a unit change. We can make the "unit-change" interpretation more meaningful by multiplying the danceability index by $100$. This linear transformation does not affect the model fit or the p-values.

```{r message=FALSE, warning=FALSE}
#Re-scale independet variable
chart_data$danceability_100 <- chart_data$danceability*100 
#Run the regression model
logit_model <- glm(top10 ~ danceability_100,family=binomial(link='logit'),data=chart_data)
#Inspect model summary
summary(logit_model )
#Inspect Pseudo R2s
logisticPseudoR2s(logit_model )
#Convert coefficients to odds ratios
exp(coef(logit_model ))
```

We observe that danceability positively affects the likelihood of becoming at top-10 hit. To get the confidence intervals for the coefficients we can use the same function as with OLS

```{r message=FALSE, warning=FALSE}
confint(logit_model)
```

In order to get a rough idea about the magnitude of the effects we can calculate the partial effects at the mean of the data (that is the effect for the average observation). Alternatively, we can calculate the mean of the effects (that is the average of the individual effects). Both can be done with the ```logitmfx(...)``` function from the ```mfx``` package. If we set ```logitmfx(logit_model, data = my_data, atmean = FALSE)``` we calculate the latter. Setting ```atmean = TRUE``` will calculate the former. However, in general we are most interested in the sign and significance of the coefficient.

```{r, message = FALSE, warning = FALSE}
library(mfx)
# Average partial effect
logitmfx(logit_model, data = chart_data, atmean = FALSE)
```

This now gives the average partial effects in percentage points. An additional point on the danceability scale (from $1$ to $100$), on average, makes it $1.57%$ more likely for a song to become at top-10 hit.

To get the effect of an additional point at a specific value, we can calculate the odds ratio by predicting the probability at a value and at the value $+1$. For example if we are interested in how much more likely a song with 51 compared to 50 danceability is to become a hit we can simply calculate the following

```{r}
#Probability of a top 10 hit with a danceability of 50
prob_50 <- exp(-(-summary(logit_model)$coefficients[1,1]-summary(logit_model)$coefficients[2,1]*50 ))
prob_50

#Probability of a top 10 hit with a danceability of 51
prob_51 <- exp(-(-summary(logit_model)$coefficients[1,1]-summary(logit_model)$coefficients[2,1]*51 ))
prob_51

#Odds ratio
prob_51/prob_50
```

So the odds are 20% higher at 51 than at 50. 

#### Logistic model with multiple predictors

Of course we can also use multiple predictors in logistic regression as shown in the formula above. We might want to add spotify followers (in million) and weeks since the release of the song.
```{r}
chart_data$spotify_followers_m <- chart_data$spotifyFollowers/1000000
chart_data$weeks_since_release <- chart_data$daysSinceRelease/7
```

Again, the familiar formula interface can be used with the ```glm()``` function. All the model summaries shown above still work with multiple predictors.

```{r}
multiple_logit_model <- glm(top10 ~ danceability_100 + spotify_followers_m + weeks_since_release,family=binomial(link='logit'),data=chart_data)
summary(multiple_logit_model)
logisticPseudoR2s(multiple_logit_model)
exp(coef(multiple_logit_model))
confint(multiple_logit_model)
```


#### Model selection

The question remains, whether a variable *should* be added to the model. We will present two methods for model selection for logistic regression. The first is based on the _Akaike Information Criterium_ (AIC). It is reported with the summary output for logit models. The value of the AIC is __relative__, meaning that it has no interpretation by itself. However, it can be used to compare and select models. The model with the lowest AIC value is the one that should be chosen. Note that the AIC does not indicate how well the model fits the data, but is merely used to compare models. 

For example, consider the following model, where we exclude the ```followers``` covariate. Seeing as it was able to contribute significantly to the explanatory power of the model, the AIC increases, indicating that the model including ```followers``` is better suited to explain the data. We always want the lowest possible AIC. 

```{r, message = FALSE, warning = FALSE}
multiple_logit_model2 <- glm(top10 ~ danceability_100 + weeks_since_release,family=binomial(link='logit'),data=chart_data)

summary(multiple_logit_model2)

```

As a second measure for variable selection, you can use the pseudo $R^2$s as shown above. The fit is distinctly worse according to all three values presented here, when excluding the spotify followers. 

```{r}
logisticPseudoR2s(multiple_logit_model2)
```


#### Predictions

We can predict the probability given an observation using the ```predict(my_logit, newdata = ..., type = "response")``` function. Replace ```...``` with the observed values for which you would like to predict the outcome variable.


```{r, message = FALSE, warning = FALSE}
# Prediction for one observation
predict(multiple_logit_model, newdata = data.frame(danceability_100=50, spotify_followers_m=10, weeks_since_release=1), type = "response")
```

The prediction indicates that a song with danceability of $50$ from an artist with $10M$ spotify followers has a $66%$ chance of being in the top-10, 1 week after its release. 

#### Perfect Prediction Logit

Perfect prediction occurs whenever a linear function of $X$ can perfectly separate the $1$s from the $0$s in the dependent variable. This is problematic when estimating a logit model as it will result in biased estimators (also check to p-values in the example!). R will return the following message if this occurs:

```glm.fit: fitted probabilities numerically 0 or 1 occurred```

Given this error, one should not use the output of the ```glm(...)``` function for the analysis. There are [various ways](https://stats.stackexchange.com/a/68917) to deal with this problem, one of which is to use Firth's bias-reduced penalized-likelihood logistic regression with the ```logistf(Y~X)``` function in the ```logistf``` package.  

##### Example

In this example data $Y = 0$ if $x_1 <0$ and $Y=1$ if $x_1>0$ and we thus have perfect prediction. As we can see the output of the regular logit model is not interpretable. The standard errors are huge compared to the coefficients and thus the p-values are $1$ despite $x_1$ being a predictor of $Y$. Thus, we turn to the penalized-likelihood version. This model correctly indicates that $x_1$ is in fact a predictor for $Y$ as the coefficient is significant.  

```{r, message = FALSE, warning = FALSE}
Y <- c(0,0,0,0,1,1,1,1)
X <- cbind(c(-1,-2,-3,-3,5,6,10,11),c(3,2,-1,-1,2,4,1,0))

# Perfect prediction with regular logit
summary(glm(Y~X, family=binomial(link="logit")))

library(logistf)
# Perfect prediction with penalized-likelihood logit
summary(logistf(Y~X))
```


## Summary {-}

* Correlation: it helps us to determine the degree to which the variation in one variable, X, is related to the variation in another variable, Y. **Note:** correlation coefficients give no indication of the direction of causality. Moreover, consideration of the third-variable problem is necessary when interpreting correlation as there might be other unobserved variables that influence the result.

* Correlation (Pearson’s) can be observed between 2 or more continuous variables (interval or ratio). If we apply Pearson’s equation on ordinal data, which we ranked beforehand, we will obtain Spearman’s correlation coefficient.

* A raw measure of the relationship between variables is the covariance. The standardized covariance (divided by the product of the two standard deviations) results in Pearson’s correlation coefficient, $r$. The correlation coefficient can be between −1 and +1. +1 indicates a perfectly positive relationship; −1 indicates a perfectly negative relationship; 0 indicates no linear relationship at all.

* The correlation coefficient is a commonly used measure of the size of an effect: values of +/-0.1 are considered a small effect, +/-0.3 a medium effect and +/-0.5 a large effect. However, an interpretation of the size of correlation should be made based on the context of the research rather than on following these benchmarks.

*	Regression analysis is a way of predicting an outcome (dependent) variable from one predictor (independent) variable (simple regression) or several predictor variables (multiple regression).

*	In a simple linear regression, we predict a dependent variable this by fitting a statistical model to the data in the form of a straight line. A regression line is a line that goes as close as possible to as many of the data points as possible.

*	R-squared indicates how much variance is explained by the model compared to how much variance there is to explain at all. It can be defined as the proportion of variance in the outcome variable that is shared by the predictor variable.

*	F-statistic denotes the extent to which the model can explain variability relative to the extent to which the model cannot explain it.

*	Beta coefficient in a single linear regression determines the slope of the regression line and the strength of the relationship between a predictor and the outcome. The coefficient is significant if Pr(>|t|) < .05. It shows the extent to which the outcome variable changes when comes to one-unit change in the predictor variable. Confidence interval of beta coefficient tell us, if we’d collected 100 samples, and calculated the confidence intervals for a beta coefficient, that 95% of these confidence intervals (out of 100) would contain the true value of beta coefficient.

*	Intercept shows the value that dependent variable holds when beta coefficients are 0.

* Assessing the regression model:

    + Check for the outliers and influential cases as they can cause a bias, because they influence the values of the estimated regression  coefficients.   
    + Check for the linearity, meaning that the relationship between the dependent and the independent variable can be reasonably approximated in linear terms.  
    + Check that the error terms have a constant variance (homoscedasticity).
    + Check that the error term is normally distributed.
    + Check for the “third-variables” that are left out the model, but influence the outcome, and therefore introduce a bias in other variables’ coefficients.
    
* In a regression model we can use categrical predictors with 2 or more categories. These predictors are included in the model as “dummy” variables. Dummy variables are factor variables that can only take two values. 

$$\begin{equation} 
x =
  \begin{cases}
    1       & \quad \text{if } i \text{th artist is international}\\
    0  & \quad \text{if } i \text{th artist is local}
  \end{cases}
\end{equation}$$

In case of multiple categorical predictors, one dummy variable cannot represent all possible values, since there are three genres (i.e., 1 = Rock, 2 = Pop, 3 = Electronic). Thus, we need to create additional dummy variables.

$$\begin{equation} 
x_5 =
  \begin{cases}
    1       & \quad \text{if } i \text{th  product is from Pop genre}\\
    0  & \quad \text{if } i \text{th product is from Rock genre}
  \end{cases}
\end{equation}$$ 

$$\begin{equation} 
x_6 =
  \begin{cases}
    1       & \quad \text{if } i \text{th  product is from Electronic genre}\\
    0  & \quad \text{if } i \text{th product is from Rock genre}
  \end{cases}
\end{equation}$$

* The standard linear regression model assumes that the relationships between the response and predictor variable is additive and linear. The additive assumption states that the effect of an independent variable on the dependent variable is independent of the values of the other independent variables included in the model. The additive assumption states that the effect of an independent variable on the dependent variable is independent of the values of the other independent variables included in the model.

* Regarding the additive assumption, it might be argued in reality that the effect of some variables are not fully independent of the values of other variables. In our example, one could argue that the effect of advertising depends on the type of artist. We can incorporate this **interaction effect** by including an interaction term in the regression equation. Interaction effect can be possible for different combinations of variable types. 

* There are cases when a linear specification might not be the best model for this data set. In this chapter we saw that a multiplicative model might be a better representation of the data we had. Multiplicative model means taking logarithms of both sides of the equation. A linear specification with log-transformed variables have important implications on interpreteation of coefficinets:  A 1% increase in independent varibale leads to a XY% increase in dependent variable.

* If our data shows initially diminishing and eventually negative returns (inverted U-shape), we can use **squared terms** to obtain a model that fits good. This results still in a linear model since the outcome variable is still explained by a linear combination of regressors even though one of the regressors is now just a non-linear function of the same variable (i.e. the squared value).

* Via linear regression models we can predict continuous outcomes such as sales or price. However, if we would like to predict a binary outcome, i.e. when the variable we want to model can only take two values  (such as 0 or 1; yes or no). For such cases when we estimate how our predictor variables change the probability of a value being 0 or 1 we use **logsitic regression.** While linear probability model produces probabilities that are above 1 and below 0 the logistic model stays between 0 and 1.

* Interpretation of coefficients in the logistic regression is a bit specific. It gives the change in the log odds of the dependent variable due to a unit change in the independent variable. This makes the exact interpretation of the coefficients difficult, but we can still interpret the signs and the p-values which will tell us if a variable has a significant positive or negative impact on the probability of the dependent variable being 1.

* Logistic regression model can consist of 1 or more independent variables. In order to determine which variables should be included, we take a look at AIC (Akaike Information Criterium) and seek for the model with the lowest AIC value. Alternatively, pseudo $R^2$ can be used as an indicator for the model fit too.

## References {-}

* Field, A., Miles J., & Field, Z. (2012): Discovering Statistics Using R. Sage Publications (**chapters 6, 7, 8**).
* James, G., Witten, D., Hastie, T., & Tibshirani, R. (2013): An Introduction to Statistical Learning with Applications in R, Springer (**chapter 3**)