---
title: "Assignment 0 (Basics)"
author: "Lacey Goodstudent - ID: 123456"
output: html_document
---
This document is for demonstration purposes only. Comments are written in *italics* and should not be included in your homework files. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Assignment

*This section is provided to you*

The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

The following variables are included in the dataset:

  * mpg: Miles/(US) gallon
  
  * cyl: Number of cylinders
  
  * disp: Displacement (cu.in.)
  
  * hp: Gross horsepower
  
  * drat: Rear axle ratio
  
  * wt: Weight (1000 lbs)
  
  * qsec: 1/4 mile time
  
  * vs: V/S
  
  * am: Transmission (0 = automatic, 1 = manual)
  
  * gear: Number of forward gears
  
  * carb: Number of carburetors

Use R and appropriate analytical techniques to answer the following questions:

1. How are miles per gallon and weight related?

2. ...

3. ... 

## Data Analysis
*This section needs to be edited by you*

### Load the data

```{r}
carData <- read.table("https://gist.githubusercontent.com/seankross/a412dfbd88b3db70b74b/raw/5f23f993cd87c283ce766e7ac6b329ee7cc2e1d1/mtcars.csv",
                      sep = ',', # make sure to use the appropriate separator!
                      stringsAsFactors = FALSE, # Do not convert string variables to factors
                      header = TRUE) # First line includes variable names
# Make sure this worked:
head(carData)
```

### Load the necessary libraries

*You can include options for single chunks in ````{r, options...}` at the beginning of each chunk. When loading the libraries it is good to set `{r, warning=FALSE, message=FALSE}` to avoid printing long messages.*

With ````{r, warning=FALSE, message=FALSE}`:

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
```

With ````{r}`:

```{r, include=FALSE, echo = FALSE}
detach(name = package:ggplot2, unload = T)
detach(package:dplyr, unload = T)
```

```{r}
library(ggplot2)
library(dplyr)
```


### Question 1

*Description of the steps before the code*

Firs we want to analyze the effect of weight on fuel efficiency...
```{r}
ggplot(carData, aes(x = wt, y = mpg)) +
  geom_line() +
  ggtitle("MPG by WT")

reg <- lm(mpg~wt, data = carData)
summary(reg)
```

*Interpretation of results after the code*

An additional 1000 pounds significantly decrease the miles per gallon by 5.35, on average. ...


### Question 2

*You may separate your code into multiple chunks per question if you prefere doing so*

#### 2.1

*Description of steps*

Next I will look at....
```{r}
ggplot(carData, aes(y = disp, x = factor(vs), group = vs)) + 
  geom_boxplot()
```

*Interpretation*

This shows that...

#### 2.2

*Description*

```{r}
t.test(disp~vs, data = carData)
```

*Interpretation*

...

## Question 3

... 

