---
title: "Assignment 3 (ANOVA)"
author: "[your name and student ID here]"
output:
  html_document: default
  pdf_document: default
---

## Assignment

The data file contains customer information from an online fashion shop. In an experiment, the customers were exposed to different types of online advertising over the past year (randomly assigned) and now you wish to analyze the results.

The following variables are included in the data set:

* customerID: unique customer ID
* revenue: revenue per cusomer for the past year (in EUR)
* gender: 0=male, 1=female
* retargeting: type of online advertising that the customer was exposed to (3 levels: 1 = no advertising, 2 = generic retargeting, 3 = dynamic retargeting)
* customerRank: ranking of customers according to their expenditure level (low rank = valuable customer, high rank = less valuable customer) 

Use R and appropriate analytical techniques to answer the following questions:

1. Has the types of online advertising an effect on revenue? Are there significant differences between the individual groups?
2. Does the effect of online advertising on revenue depend on gender? (conduct a 2 × 3 analysis of variance with "revenue" as the dependent variable and "gender" and "retargeting" as the independent variables or factors.)
3. Is the customer ranking significantly influenced by the type of online advertising? Are there significant differences between the individual groups?

When answering the questions, please remember to address the following points:

* Formulate the corresponding hypotheses and choose an appropriate statistical test
* Provide the reason for your choice and discuss if the assumptions of the test are met 
* Convert the variables to the appropriate type (e.g., factor variables)
* Create appropriate graphs to explore the data (e.g., boxplot, bar chart)
* Provide appropriate descriptive statistics for the variables
* Report and interpret the test results accurately
* Finally, don't forget to report your research conclusion in an appropriate way

When you are done with your analysis, click on "Knit to HTML" button above the code editor. This will create a HTML document of your results in the folder where the "assignment3.Rmd" file is stored. Open this file in your Internet browser to see if the output is correct. If the output is correct, submit the HTML file via Learn\@WU. The file name should be "assignment3_studendID_name.html".

## Data analysis

## Load data

```{r load_data}
customer_data <- read.table("https://raw.githubusercontent.com/IMSMWU/Teaching/master/MRDA2017/assignment3.4.dat", 
                          sep = "\t", 
                          header = TRUE) #read in data
head(customer_data)
str(customer_data)
```

## Question 1

Provide a description of your steps here!

```{r question_1_1}
# provide your code here (you can use multiple code chunks per question if you like)
# the output will automatically appear below the code chunk when you create the html file
```

Interpret the results here!

## Question 2

Provide a description of your steps here!

```{r question_2_1}
# provide your code here (you can use multiple code chunks per question if you like)
```

Interpret the results here!

## Question 3

Provide a description of your steps here!

```{r question_3_1}
# provide your code here (you can use multiple code chunks per question if you like)
```

Interpret the results here!
