---
title: "Assignment 2 (Hypothesis Testing)"
author: "[your name and student ID here]"
output:
  html_document: default
  pdf_document: default
---

## Assignment

As a marketing manager at a video streaming service, you are interested in the effect of online advertising on the number of streams that a movie receives. To test the effect of online advertising on streams, you select a representative sample of 200 movies and randomly assign 100 movies to be included in an online advertising campaign. The other half of the sample serves as the control group. You run the experiment for one week and collect data regarding the number of streams from this period. In addition, you collect other variables about the movies that you wish to analyze. Overall, the data set includes the following variables:

* movieID: unique movie ID
* streams_sd: number of streams in SD-quality
* streams_hd: number of streams in HD-quality
* online_advertising: indicator whether a movie was included in the online advertising campaign (0 = no, 1 = yes) 
* customer_rating: satisfaction measured on a 5-point Likert scale after the moviews were watched: (1) "very dissatisfied" – (5) "very satisfied""
* academy_award: 0= non-nominated film, 1=nominated film
* genre: 1 = drama, 2 = romance, 3 = comedy, 4 = thriller, 5 = action

Apply appropriate statistical methods to answer the following questions:

1. Is there a significant difference in streams between movies that were included in the online advertising campaign and those that were not included? (Please conduct the test for SD and HD movies)
2. Is there a significant difference in streams between movies in HD and SD quality?
3. Is there a significant association between the genre of a movie and the academy award nomination?
4. Is there a significant difference in customer ratings between nominated and non-nominated movies?

When answering the questions, please remember to address the following points:

* Formulate the corresponding hypotheses and choose an appropriate statistical test
* Provide the reason for your choice and discuss if the assumptions of the test are met 
* Convert the variables to the appropriate type (e.g., factor variables)
* Create appropriate graphs to explore the data (e.g., boxplot, bar chart, histogram)
* Provide appropriate descriptive statistics for the variables
* Report and interpret the test results accurately (including confidence intervals where appropriate)  
* Finally, don't forget to report your research conclusion in an appropriate way

When you are done with your analysis, click on "Knit to HTML" button above the code editor. This will create a HTML document of your results in the folder where the "assignment.Rmd" file is stored. Open this file in your Internet browser to see if the output is correct. If the output is correct, submit the HTML file via Learn\@WU. The file name should be "assignment2_studendID_name.html".

## Data analysis

## Load data

```{r load_data}
movie_data <- read.table("https://raw.githubusercontent.com/IMSMWU/Teaching/master/MRDA2017/assignment1.5.dat", 
                          sep = "\t", 
                          header = TRUE) #read in data
head(movie_data)
str(movie_data)
```

## Question 1

Provide a description of your steps here!

```{r question_1_1}
# provide your code here (you can use multiple code chunks per question if you like)
# the output will automatically appear below the code chunk when you create the html file
```

Interpret the results here!

## Question 2

Provide a description of your steps here!

```{r question_2_1}
# provide your code here (you can use multiple code chunks per question if you like)
```

Interpret the results here!

## Question 3

Provide a description of your steps here!

```{r question_3_1}
# provide your code here (you can use multiple code chunks per question if you like)
```

Interpret the results here!

## Question 4

Provide a description of your steps here!

```{r question_4_1}
# provide your code here (you can use multiple code chunks per question if you like)
```

Interpret the results here!
